---
title: "dimensionality_reduction"
format: html
editor: source
---

```{r, message=FALSE}
library(tidyverse)
library(ggbiplot)
library(glue)

source("../../R/load_functions.R")
triggerR <- load_trigger_functions()
rm(load_trigger_functions)

dir.create("../../outputs/dimensionality_reduction", showWarnings = F)
```

```{r}
if (!exists("pacmap")){
  reticulate::use_condaenv("pacmap", required = TRUE)
  pacmap <- reticulate::import("pacmap")
}
```

## Get Data

```{r}
daily_phy <- triggerR$make_daily_wide_table(
  domain_filter = "physiological", 
  return_meta = T, with_sleep = T,
  exclude_measurements = c("oxygens", "sleeprate", "breathrate", "insomnia"))

daily_phy_nosleep <- triggerR$make_daily_wide_table(
  domain_filter = "physiological", 
  return_meta = T, with_sleep = F,
  exclude_measurements = c("oxygens", "sleeprate", "breathrate", "insomnia"))

daily_env <- triggerR$make_daily_wide_table(
  domain_filter = "environmental", 
  return_meta = T, with_sleep = F)

daily_env_nopollution <- triggerR$make_daily_wide_table(
  domain_filter = "environmental", 
  return_meta = T, with_sleep = F,
  exclude_measurements = c("pm1", "pm25", "pm10", "pc03", "pc05", "pc1", "pc25", "pc5", "pc10"))
```

## PCA

```{r}
pca_phy <- prcomp(daily_phy$data, scale. = TRUE)
pca_env <- prcomp(daily_env$data, scale. = TRUE)
pca_phy_nosleep <- prcomp(daily_phy_nosleep$data, scale. = TRUE)
pca_env_pollution <- prcomp(daily_env_nopollution$data, scale. = TRUE)
```

#### PCA phy sleep

```{r, fig.width=9, fig.height=9, message=F}
p_biplot_pca_phy <- ggbiplot(pca_phy,
         var.factor = 2, point.size = 1, alpha = 0.1, var.axes = TRUE, 
         varname.size = 2.5, varname.adjust = 3, varname.color = "steelblue4",
         circle = TRUE) +
  labs(title = "PCA biplot - Physiological triggers with sleep") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.25, color = "grey90")) +
  coord_fixed(clip = "off", xlim = c(-4,4), ylim = c(-4,4))
p_biplot_pca_phy

ggsave("../../outputs/dimensionality_reduction/pca_biplot_phy.png", p_biplot_pca_phy,
       height = 9, width = 9, dpi = 300)
```

#### PCA phy no sleep

```{r, fig.width=9, fig.height=9, message=F}
p_biplot_pca_phy_nosleep <- ggbiplot(pca_phy_nosleep,
         var.factor = 2, point.size = 1, alpha = 0.1, var.axes = TRUE, 
         varname.size = 2.5, varname.adjust = 3, varname.color = "steelblue4",
         circle = TRUE) +
  labs(title = "PCA biplot - Physiological triggers without sleep") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.25, color = "grey90")) +
  coord_fixed(clip = "off", xlim = c(-4,4), ylim = c(-4,4))

p_biplot_pca_phy_nosleep

ggsave("../../outputs/dimensionality_reduction/pca_biplot_phy_nosleep.png", p_biplot_pca_phy_nosleep,
       height = 9, width = 9, dpi = 300)
```

#### PCA env

```{r, fig.width=9, fig.height=9, message=F}
p_biplot_pca_env <- ggbiplot(pca_env,
         var.factor = 2, point.size = 1, alpha = 0.1, var.axes = TRUE, 
         varname.size = 2.5, varname.adjust = 3, varname.color = "steelblue4",
         circle = TRUE) +
  labs(title = "PCA biplot - Environmental triggers") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.25, color = "grey90")) +
  coord_fixed(clip = "off", xlim = c(-4,4), ylim = c(-4,4))

p_biplot_pca_env

ggsave("../../outputs/dimensionality_reduction/pca_biplot_env.png", p_biplot_pca_env,
       height = 9, width = 9, dpi = 300)
```

#### PCA phy no pollution

```{r, fig.width=9, fig.height=9, message=F}
pca_biplot_env_nopollution <- ggbiplot(pca_env_pollution,
         var.factor = 2, point.size = 1, alpha = 0.1, var.axes = TRUE, 
         varname.size = 2.5, varname.adjust = 3, varname.color = "steelblue4",
         circle = TRUE) +
  labs(title = "PCA biplot - Environmental triggers nopollution") +
  theme_minimal(base_size = 13) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major = element_line(linewidth = 0.25, color = "grey90")) +
  coord_fixed(clip = "off", xlim = c(-4,4), ylim = c(-4,4))

pca_biplot_env_nopollution

ggsave("../../outputs/dimensionality_reduction/pca_biplot_env_nopollution.png", pca_biplot_env_nopollution,
       height = 9, width = 9, dpi = 300)
```

## PACMAP

```{r}
reducer_env <- pacmap$PaCMAP(
  n_components = 2L,
  n_neighbors = 15L,
  #distance = "manhattan",
  random_state = 42L,
  verbose = FALSE)

reducer_phy <- pacmap$PaCMAP(
  n_components = 2L,
  n_neighbors = 15L,
  #distance = "manhattan",
  random_state = 42L,
  verbose = FALSE)
```

#### PACMAP ENV

```{r}
pacmap_env <- reducer_env$fit_transform(scale(daily_env$data))
```

```{r, fig.width=8, fig.height=8, warning=FALSE}
p_pacmap_env <- as_tibble(pacmap_env) |>
  setNames(c("PaCMAP1", "PaCMAP2")) %>%
  mutate(event = rownames(daily_env$data), .before = 1) %>%
  left_join(as_tibble(daily_env$meta, rownames = "event"), by = "event") %>%
  rename(nationality = country) %>%
  ggplot(aes(x = PaCMAP1, y = PaCMAP2, color = nationality)) +
  geom_point(alpha = .5) +
  theme_minimal(base_size = 13) +
  coord_fixed() +
  theme(legend.position = "top") +
  labs(title = "PaCMAP - Environmental trigger")

p_pacmap_env

ggsave("../../outputs/pacmap_env.png", p_pacmap_env,
       width = 8, height = 8, dpi = 300)
```

#### PACMAP PHY

```{r}
numerous_user <- daily_phy$meta %>%
  rownames_to_column("event") %>%
  count(userId, name = "n_events") |>
  filter(n_events >= 150) |>
  pull(userId)

numerous_user_events <- daily_phy$meta %>%
  as_tibble(rownames = "event") %>%
  filter(userId %in% numerous_user) %>%
  pull(event)

daily_phy_sub <- daily_phy
daily_phy_sub$data <- daily_phy$data[numerous_user_events, ]
daily_phy_sub$meta <- daily_phy$meta[numerous_user_events, ]
```

```{r}
pacmap_phy <- reducer_phy$fit_transform(scale(daily_phy_sub$data)) %>%
  as_tibble(pacmap_phy) |>
  setNames(c("PaCMAP1", "PaCMAP2")) %>%
  mutate(event = rownames(daily_phy_sub$data), .before = 1) %>%
  left_join(as_tibble(daily_phy$meta, rownames = "event"), by = "event") 
```

```{r, fig.width=8, fig.height=8}
p_pacmap_phy <- pacmap_phy |>
  ggplot(aes(x = PaCMAP1, y = PaCMAP2, color = userId)) +
  geom_point(alpha = .8) +
  ggqualpal::scale_color_qualpal(colorspace = list(h = c(0, 360),
                                                   s = c(0.5, 0.7),
                                                   l = c(0.6, 0.8))) +
  theme_minimal(base_size = 13) +
  coord_fixed() +
  theme(legend.position = "top") +
  labs(title = "PaCMAP - Environmental trigger") +
  guides(color = guide_legend(nrow = 1))

p_pacmap_phy

ggsave("../../outputs/pacmap_phy.png", p_pacmap_phy,
       width = 8, height = 8, dpi = 300)
```

```{r}
phy_adonis <- vegan::adonis2(dist(scale(daily_phy_sub$data)) ~ userId,
               data = daily_phy_sub$meta, permutations = 999)
phy_adonis
```

```{r}
phy_disp <- vegan::betadisper(dist(scale(daily_phy_sub$data)),
                                   group = daily_phy_sub$meta$userId)
```

```{r}
phy_anova <- anova(phy_disp)
phy_anova
```

```{r}
phy_perm <- vegan::permutest(phy_disp, permutations = 999)
phy_perm
```

##### Note on subject-specific physiological profiles

The multivariate analysis of daily physiological profiles suggests a clear subject-specific structure. Using Euclidean distance on scaled physiological measurements, PERMANOVA (`vegan::adonis2`) showed a significant effect of `userId` on the overall multivariate profiles (*R²* = 0.242, *F* = 69.942, *p* = 0.001), indicating that subjects differ in their physiological patterns.

However, the test for homogeneity of multivariate dispersion (`vegan::betadisper`, followed by `anova` and `permutest`) was also significant (*F* = 24.894, *p* = 0.001). This indicates that the observed subject-specific structure is not only due to differences in the position of users in the multivariate space, but also to differences in within-subject daily variability. In other words, some users show more stable physiological profiles across days, whereas others display greater day-to-day variability.

Therefore, the `userId` effect should be interpreted as a combination of differences among subjects’ average multivariate physiological profiles and differences in their intra-subject dispersion.

#### Silhouette score on physiological

```{r}
phy_sil <- cluster::silhouette(
  x = as.integer(factor(daily_phy_sub$meta$userId)),
  dist = dist(scale(daily_phy_sub$data)))

mean(phy_sil[, "sil_width"])
```

Although PERMANOVA showed a significant effect of `userId` on scaled physiological profiles using Euclidean distance (`vegan::adonis2`: R² = 0.242, F = 69.942, p = 0.001), the average silhouette width was close to zero (mean silhouette = 0.020). This indicates that users do not form clearly separated and compact clusters in the original multivariate space. Moreover, the significant dispersion test (`vegan::betadisper`: F = 24.894, p = 0.001) suggests that the subject-specific structure is partly associated with differences in within-user variability. Overall, `userId` has a statistically significant effect on physiological profiles, but the separation among users should be interpreted as partial rather than as well-defined clustering.









