---
title: "Hourly wearable data example - Italy / Bologna"
subtitle: "From long raw hourly data to an easier-to-use table"
format: html
editor: source
---

```{r, message=FALSE}
library(tidyverse)
```

# Aim of the notebook

This notebook shows how to work with the hourly wearable dataset prepared for the Italian/Bologna participants.

The input file contains wearable measurements aggregated at the **hourly level** and stored in a **long format**.  
Each row corresponds mainly to a specific combination of:

- participant (`userId`);
- date;
- hour;
- wearable measurement (`measurement`);
- summary statistic type (`type`);
- aggregated value (`value`);
- number of valid observations used to compute the aggregated value (`n_obs_valid`).

These are the main variables needed to interpret the dataset, but they are not the only columns available in the table.

In particular, `n_obs_valid` indicates how many valid observations were used to compute each hourly aggregated value.  
Because it quantifies the amount of data supporting each aggregation, it is the main variable that can be used to filter the dataset and assess whether a given hourly measurement is reliable enough to be included in the analysis.

In other words, `n_obs_valid` tells us how many underlying measurements contributed to the hourly `mean`, `min`, `max`, or `sum`.
Most wearable measurements were aggregated using hourly `mean`, `min`, and `max`, while `sum` was used only for cumulative variables such as calories and steps.

## Load Data

```{r}
wearable_long_italy_hourly <- read_tsv("data/wearable_long_table_raw.tsv.gz",
                                       show_col_types = FALSE) %>%
  mutate(userId = as.factor(userId),
         date = as_date(date))
```

```{r}
glimpse(wearable_long_italy_hourly)
```

```{r}
wearable_long_italy_hourly %>%
  summarise(
    n_rows = n(),
    n_subjects = n_distinct(userId),
    first_date = min(date, na.rm = TRUE),
    last_date = max(date, na.rm = TRUE)) %>%
  gt::gt()
```

# Filter the long table

In this step we create a simplified long-format table focused on the wearable variables that are more directly useful for physiological, activity-related, and environmental analyses.

We apply three main filters.

First, GPS data are excluded.  
This choice was made because the aim of this example is to focus on physiological and wearable-derived variables rather than on movement trajectories or spatial information. GPS data can still be included in a separate analysis, but they may require specific preprocessing and filtering criteria, for example to account for location accuracy, duplicated positions, or irregular sampling.

Second, we retain only `mean` and `sum` hourly aggregations.  
At the hourly level, `mean` is used as the main summary statistic for most wearable measurements, while `sum` is retained for cumulative variables such as steps and calories. If needed, hourly `min` and `max` values can also be included to describe within-hour variability or to derive additional indicators, such as hourly ranges.

Third, we keep only hourly aggregated values based on at least 10 valid observations.  
This threshold is not intended as a fixed rule, but as a practical compromise for this example. The idea is to remove hourly values computed from too few observations, while retaining a reasonable amount of data for exploratory analyses.
This choice should also be interpreted in light of the different sampling frequencies of the wearable devices. Some smartwatch-derived variables are sampled approximately every minute, while others may be recorded less frequently or with a less regular interval. For this reason, the threshold used here is only a working criterion and can be revised depending on the variable and on the desired level of reliability.

The `n_obs_valid` threshold can be changed depending on the specific measurement, the expected sampling frequency of the device, and the level of reliability required for the analysis.

```{r, warning=FALSE, fig.width=12, fig.height=4}
wearable_long_italy_hourly %>%
  ggplot(aes(x = n_records_total, fill = source)) +
  geom_histogram(bins = 10) +
  facet_wrap(~measurement, ncol = 7) +
  theme_bw() +
  coord_cartesian(xlim = c(0, 65)) +
  theme(legend.position = "top") +
  labs(x = "Total recorded events per subject-hour",
       y = "Count",
       fill = "Source")
```

```{r}
long_filt <- wearable_long_italy_hourly %>%
  filter(source != "gps") %>%
  filter(type %in% c("mean", "sum")) %>%
  select(
    userId,
    date,
    hour,
    n_records_total,
    n_obs_valid,
    measurement,
    value
  ) %>%
  filter(n_obs_valid >= 10)
```


# Convert to complete wide hourly table

The filtered table contains only the hours where at least one valid wearable measurement is available.

For many analyses, however, it is useful to have a complete hourly table, including participant-hours with no wearable data at all.

In this step, we create a complete grid of all combinations of participant, date, and hour, and then join the available wearable measurements in wide format.

The resulting table contains one row for each participant-hour.  
Hours without any valid wearable measurement are retained as rows with `NA` values across all measurement columns.

```{r}
wide <- long_filt %>%
  distinct(userId) %>%
  crossing(date = seq(min(long_filt$date, na.rm = TRUE),
                      max(long_filt$date, na.rm = TRUE),
                      by = "day"),
           hour = 0:23) %>%
  unite(col = "event", userId, date, hour, sep = "--", remove = FALSE) %>%
  left_join(long_filt %>%
              unite(col = "event", userId, date, hour, sep = "--", remove = FALSE) %>%
      select(event, measurement, value) %>%
      pivot_wider(names_from = measurement, values_from = value), by = "event")
```

```{r, fig.width=16, fig.height=6}
wide %>%
  arrange(userId, date, hour) %>%
  select(-userId, -date, -hour) %>%
  pivot_longer(-event, names_to = "measurements", values_to = "value") %>%
  mutate(presence = !is.na(value)) %>%
  ggplot(aes(x = event, y = measurements, fill = presence)) +
  geom_tile(linewidth = 0) +
  theme(legend.position = "top",
        axis.text.x = element_blank()) +
  scale_fill_manual(
    values = c("FALSE" = "grey90", "TRUE" = "#2C7FB8"),
    labels = c("FALSE" = "Missing", "TRUE" = "Available"),
    name = NULL) 
```

**Figure. Availability of hourly aggregated wearable measurements for the Italian/Bologna participants.**  
Columns represent participant-hour events, where each event is defined by the combination of participant identifier (`userId`), calendar date (`date`), and hour of the day (`hour`). Rows represent wearable measurements. Blue tiles indicate available aggregated values after filtering, whereas grey tiles indicate missing values.

# Save the simplified wide table

The resulting table contains one row per participant and hour.

Rows with no available wearable measurements are kept in the table and contain `NA` values in the measurement columns.

```{r}
write_tsv(
  wide,
  "data/wearable_hourly_wide_italy_bologna.tsv.gz"
)
```

# Notes 

The final table is designed to be easier to use for integration with hourly environmental or climate variables.

The key columns are:

- `userId`: participant identifier;
- `date`: calendar date;
- `hour`: hour of the day;
- `event`: compact participant-hour identifier;
- wearable measurement columns: one column for each selected measurement.

Only non-GPS wearable measurements are included.

Only hourly `mean` and `sum` values are retained in this simplified table.  
The `mean` aggregation is used for most wearable measurements, while `sum` is used for cumulative variables such as calories and steps.

The filtering threshold `n_obs_valid >= 10` is applied before creating the final wide table.  
This means that the final table contains only hourly values based on at least 10 valid underlying observations.

Participant-hours with no reliable wearable measurements are still retained as rows, with `NA` values in the measurement columns.

```{r}
write_tsv(wide, "data/wearable_hourly_wide_italy_bologna.tsv.gz")
```
