---
title: "pytrigger exploration"
engine: knitr
---

## Environment

The notebook uses `reticulate` to run Python and R in the same document.  
This first check confirms which Python environment is active.

```{r}
library(reticulate)

py_config()
```

## Setup

Load the required Python libraries and create dedicated folders for the data
and outputs produced by this notebook.

```{python}
import ssl
import sys
from pathlib import Path

import pandas as pd
from trigger import TriggerDB

NOTEBOOK_ID = "00_pytrigger"

DATA_DIR = Path("data") / NOTEBOOK_ID
OUTPUT_DIR = Path("outputs") / NOTEBOOK_ID

DATA_DIR.mkdir(parents=True, exist_ok=True)
OUTPUT_DIR.mkdir(parents=True, exist_ok=True)

print(f"Python: {sys.executable}")
print(f"OpenSSL: {ssl.OPENSSL_VERSION}")
print(f"Working directory: {Path.cwd()}")
print(f"Data directory: {DATA_DIR.resolve()}")
print(f"Output directory: {OUTPUT_DIR.resolve()}")
```

## Available tables

Connect to the TRIGGER API and list the tables currently available through
`pytrigger`.

```{python}
with TriggerDB() as db:
    tables = db.tables()

print(f"Number of available tables: {len(tables)}")

tables
```

## Available columns

Inspect the columns exposed for a selected table before building a query.

```{python}
with TriggerDB() as db:
    myair_hourly_columns = db.columns("myair_hourly")

myair_hourly_columns
```

## Select data

`select()` retrieves a filtered subset of a table.  
Here, the first 1,000 observations for user 187 are ordered by timestamp.

```{python}
with TriggerDB() as db:
    myair_tidy_get = pd.DataFrame(
        db.select(
            table="myair_tidy",
            columns=[
                "userId",
                "event_ts",
                "pm25",
                "temperature",
                "humidity",
            ],
            where={
                "userId": "=187",
            },
            order_by="event_ts",
            order="ASC",
            limit=1000,
        )
    )

myair_tidy_r = {
    column: myair_tidy_get[column].astype(object).tolist()
    for column in myair_tidy_get.columns
}

print(f"Shape: {myair_tidy_get.shape}")

myair_tidy_get.head()
```

## Plot data in R

The selected columns are converted to simple Python lists so that `reticulate`
can transfer them safely to R.

```{r}
library(tidyverse)

py$myair_tidy_r %>%
  as_tibble() %>%
  mutate(
    userId = as.character(userId),
    event_ts = as.POSIXct(event_ts)
  ) %>%
  ggplot(aes(x = event_ts, y = pm25)) +
  geom_point() +
  labs(
    title = "PM2.5 over time",
    x = NULL,
    y = "PM2.5"
  ) +
  theme_minimal()
```

## Download complete table

`download()` retrieves the complete compressed dump and saves it in the data
folder associated with this notebook.

```{python}
dump_file = DATA_DIR / "myair_tidy.csv.gz"

with TriggerDB() as db:
    myair_tidy_dump = db.download(
        table="myair_tidy",
        outfile=dump_file,
    )

print(f"Shape: {myair_tidy_dump.shape}")
print(f"File: {dump_file.resolve()}")

myair_tidy_dump.head()
```
