---
title: "Data quality control"
author: "Alessandro Fuschi"
editor: source
toc: true
---

## Setup

```{r, message=FALSE}
library(tidyverse)

cfg <- list(
  daily_dir = file.path("data", "daily"),
  five_min_dir = file.path("data", "5min"),
  tidy_dir = file.path("data", "tidy"),
  analysis_start_date = as.Date("2025-03-01"),
  timezone = "UTC"
)
```

## Aim

This report centralises structural checks that are common to the analyses. It counts missing temporal keys, records on or before 1 March 2025, records from the current date or later, and duplicated analytical keys within the retained date window.

Daily tables use `userId` and `date` as their analytical key. Five-minute tables use `userId` and `bucket_5min`. The duplicate summary reports only aggregate counts and never displays participant identifiers. All rows belonging to duplicated keys are excluded from downstream analyses when their origin cannot be resolved unambiguously.

These checks describe table structure rather than the validity of individual measurements. The report reads local data without modifying it.

## Daily and sleep tables

Only the columns needed to inspect analytical keys are imported.

```{r, message=FALSE}
daily_keys <- bind_rows(
  read_csv(
    file.path(cfg$daily_dir, "gps_daily.csv.gz"), col_select = c(userId, date),
    col_types = cols(userId = col_character(), date = col_date()),
    na = c("", "NA", "NULL"), show_col_types = FALSE
  ) %>% mutate(data_table = "gps_daily", resolution = "Daily"),
  read_csv(
    file.path(cfg$daily_dir, "myair_daily.csv.gz"), col_select = c(userId, date),
    col_types = cols(userId = col_character(), date = col_date()),
    na = c("", "NA", "NULL"), show_col_types = FALSE
  ) %>% mutate(data_table = "myair_daily", resolution = "Daily"),
  read_csv(
    file.path(cfg$daily_dir, "smartwatchhigh_daily.csv.gz"), col_select = c(userId, date),
    col_types = cols(userId = col_character(), date = col_date()),
    na = c("", "NA", "NULL"), show_col_types = FALSE
  ) %>% mutate(data_table = "smartwatchhigh_daily", resolution = "Daily"),
  read_csv(
    file.path(cfg$daily_dir, "smartwatchlow_daily.csv.gz"), col_select = c(userId, date),
    col_types = cols(userId = col_character(), date = col_date()),
    na = c("", "NA", "NULL"), show_col_types = FALSE
  ) %>% mutate(data_table = "smartwatchlow_daily", resolution = "Daily"),
  read_csv(
    file.path(cfg$tidy_dir, "sleep_tidy.csv.gz"), col_select = c(userId, date),
    col_types = cols(userId = col_character(), date = col_date()),
    na = c("", "NA", "NULL"), show_col_types = FALSE
  ) %>% mutate(data_table = "sleep_tidy", resolution = "Participant-night")
)

daily_date_summary <- daily_keys %>%
  group_by(data_table, resolution) %>%
  summarise(
    rows = n(), missing_user_ids = sum(is.na(userId)), missing_dates = sum(is.na(date)),
    on_or_before_start = sum(date <= cfg$analysis_start_date, na.rm = TRUE),
    within_window = sum(date > cfg$analysis_start_date & date < Sys.Date(), na.rm = TRUE),
    today_or_later = sum(date >= Sys.Date(), na.rm = TRUE), .groups = "drop"
  )

daily_retained_rows <- daily_keys %>%
  filter(!is.na(userId), !is.na(date), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  count(data_table, resolution, name = "retained_rows")

daily_duplicate_summary <- daily_keys %>%
  filter(!is.na(userId), !is.na(date), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  count(data_table, resolution, userId, date, name = "key_n") %>%
  filter(key_n > 1) %>%
  group_by(data_table, resolution) %>%
  summarise(duplicated_keys = n(), rows_involved = sum(key_n), .groups = "drop") %>%
  right_join(daily_retained_rows, by = c("data_table", "resolution"), relationship = "one-to-one") %>%
  mutate(
    duplicated_keys = replace_na(duplicated_keys, 0L), rows_involved = replace_na(rows_involved, 0L),
    percent_rows_involved = 100 * rows_involved / retained_rows
  )

rm(daily_keys, daily_retained_rows)
```

## Five-minute tables

The same checks are repeated on the canonical five-minute tables. Timestamps are interpreted in UTC, consistently with the high-resolution analyses. Tables are processed one at a time to keep memory use limited.

```{r, message=FALSE}
gps_5min <- read_csv(
  file.path(cfg$five_min_dir, "gps_5min.csv.gz"), col_select = c(userId, bucket_5min),
  col_types = cols(userId = col_character(), bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")),
  locale = locale(tz = cfg$timezone), na = c("", "NA", "NULL"), show_col_types = FALSE
) %>% mutate(date = as.Date(bucket_5min, tz = cfg$timezone))

gps_5min_date_summary <- gps_5min %>%
  summarise(
    data_table = "gps_5min", resolution = "5 minutes", rows = n(),
    missing_user_ids = sum(is.na(userId)), missing_dates = sum(is.na(bucket_5min)),
    on_or_before_start = sum(date <= cfg$analysis_start_date, na.rm = TRUE),
    within_window = sum(date > cfg$analysis_start_date & date < Sys.Date(), na.rm = TRUE),
    today_or_later = sum(date >= Sys.Date(), na.rm = TRUE)
  )

gps_5min_retained_rows <- gps_5min %>%
  filter(!is.na(userId), !is.na(bucket_5min), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  summarise(retained_rows = n()) %>% pull()

gps_5min_duplicate_summary <- gps_5min %>%
  filter(!is.na(userId), !is.na(bucket_5min), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n > 1) %>%
  summarise(duplicated_keys = n(), rows_involved = sum(key_n)) %>%
  mutate(
    data_table = "gps_5min", resolution = "5 minutes", retained_rows = gps_5min_retained_rows,
    percent_rows_involved = 100 * rows_involved / retained_rows
  ) %>%
  select(data_table, resolution, duplicated_keys, rows_involved, retained_rows, percent_rows_involved)

rm(gps_5min, gps_5min_retained_rows)
invisible(gc())
```

```{r, message=FALSE}
myair_5min <- read_csv(
  file.path(cfg$five_min_dir, "myair_5min.csv.gz"), col_select = c(userId, bucket_5min),
  col_types = cols(userId = col_character(), bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")),
  locale = locale(tz = cfg$timezone), na = c("", "NA", "NULL"), show_col_types = FALSE
) %>% mutate(date = as.Date(bucket_5min, tz = cfg$timezone))

myair_5min_date_summary <- myair_5min %>%
  summarise(
    data_table = "myair_5min", resolution = "5 minutes", rows = n(),
    missing_user_ids = sum(is.na(userId)), missing_dates = sum(is.na(bucket_5min)),
    on_or_before_start = sum(date <= cfg$analysis_start_date, na.rm = TRUE),
    within_window = sum(date > cfg$analysis_start_date & date < Sys.Date(), na.rm = TRUE),
    today_or_later = sum(date >= Sys.Date(), na.rm = TRUE)
  )

myair_5min_retained_rows <- myair_5min %>%
  filter(!is.na(userId), !is.na(bucket_5min), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  summarise(retained_rows = n()) %>% pull()

myair_5min_duplicate_summary <- myair_5min %>%
  filter(!is.na(userId), !is.na(bucket_5min), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n > 1) %>%
  summarise(duplicated_keys = n(), rows_involved = sum(key_n)) %>%
  mutate(
    data_table = "myair_5min", resolution = "5 minutes", retained_rows = myair_5min_retained_rows,
    percent_rows_involved = 100 * rows_involved / retained_rows
  ) %>%
  select(data_table, resolution, duplicated_keys, rows_involved, retained_rows, percent_rows_involved)

rm(myair_5min, myair_5min_retained_rows)
invisible(gc())
```

```{r, message=FALSE}
smartwatch_high_5min <- read_csv(
  file.path(cfg$five_min_dir, "smartwatchhigh_5min.csv.gz"), col_select = c(userId, bucket_5min),
  col_types = cols(userId = col_character(), bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")),
  locale = locale(tz = cfg$timezone), na = c("", "NA", "NULL"), show_col_types = FALSE
) %>% mutate(date = as.Date(bucket_5min, tz = cfg$timezone))

smartwatch_high_5min_date_summary <- smartwatch_high_5min %>%
  summarise(
    data_table = "smartwatchhigh_5min", resolution = "5 minutes", rows = n(),
    missing_user_ids = sum(is.na(userId)), missing_dates = sum(is.na(bucket_5min)),
    on_or_before_start = sum(date <= cfg$analysis_start_date, na.rm = TRUE),
    within_window = sum(date > cfg$analysis_start_date & date < Sys.Date(), na.rm = TRUE),
    today_or_later = sum(date >= Sys.Date(), na.rm = TRUE)
  )

smartwatch_high_5min_retained_rows <- smartwatch_high_5min %>%
  filter(!is.na(userId), !is.na(bucket_5min), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  summarise(retained_rows = n()) %>% pull()

smartwatch_high_5min_duplicate_summary <- smartwatch_high_5min %>%
  filter(!is.na(userId), !is.na(bucket_5min), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n > 1) %>%
  summarise(duplicated_keys = n(), rows_involved = sum(key_n)) %>%
  mutate(
    data_table = "smartwatchhigh_5min", resolution = "5 minutes", retained_rows = smartwatch_high_5min_retained_rows,
    percent_rows_involved = 100 * rows_involved / retained_rows
  ) %>%
  select(data_table, resolution, duplicated_keys, rows_involved, retained_rows, percent_rows_involved)

rm(smartwatch_high_5min, smartwatch_high_5min_retained_rows)
invisible(gc())
```

```{r, message=FALSE}
smartwatch_low_5min <- read_csv(
  file.path(cfg$five_min_dir, "smartwatchlow_5min.csv.gz"), col_select = c(userId, bucket_5min),
  col_types = cols(userId = col_character(), bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")),
  locale = locale(tz = cfg$timezone), na = c("", "NA", "NULL"), show_col_types = FALSE
) %>% mutate(date = as.Date(bucket_5min, tz = cfg$timezone))

smartwatch_low_5min_date_summary <- smartwatch_low_5min %>%
  summarise(
    data_table = "smartwatchlow_5min", resolution = "5 minutes", rows = n(),
    missing_user_ids = sum(is.na(userId)), missing_dates = sum(is.na(bucket_5min)),
    on_or_before_start = sum(date <= cfg$analysis_start_date, na.rm = TRUE),
    within_window = sum(date > cfg$analysis_start_date & date < Sys.Date(), na.rm = TRUE),
    today_or_later = sum(date >= Sys.Date(), na.rm = TRUE)
  )

smartwatch_low_5min_retained_rows <- smartwatch_low_5min %>%
  filter(!is.na(userId), !is.na(bucket_5min), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  summarise(retained_rows = n()) %>% pull()

smartwatch_low_5min_duplicate_summary <- smartwatch_low_5min %>%
  filter(!is.na(userId), !is.na(bucket_5min), date > cfg$analysis_start_date, date < Sys.Date()) %>%
  count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n > 1) %>%
  summarise(duplicated_keys = n(), rows_involved = sum(key_n)) %>%
  mutate(
    data_table = "smartwatchlow_5min", resolution = "5 minutes", retained_rows = smartwatch_low_5min_retained_rows,
    percent_rows_involved = 100 * rows_involved / retained_rows
  ) %>%
  select(data_table, resolution, duplicated_keys, rows_involved, retained_rows, percent_rows_involved)

rm(smartwatch_low_5min, smartwatch_low_5min_retained_rows)
invisible(gc())
```

## Date-window summary

Rows in the analysis window have non-missing temporal keys and satisfy `2025-03-01 < date < Sys.Date()`. Missing participant identifiers are shown separately and are not eligible for the duplicate-key check.

```{r}
date_summary <- bind_rows(
  daily_date_summary, gps_5min_date_summary, myair_5min_date_summary,
  smartwatch_high_5min_date_summary, smartwatch_low_5min_date_summary
) %>% arrange(factor(resolution, levels = c("5 minutes", "Daily", "Participant-night")), data_table)

date_summary %>%
  knitr::kable(
    format = "pipe", big.mark = ",",
    col.names = c(
      "Data table", "Resolution", "Rows", "Missing user IDs", "Missing dates/timestamps",
      "On or before start", "Within window", "Today or later"
    ),
    caption = paste0("Date-window checks; retained window: ", cfg$analysis_start_date, " < date < ", Sys.Date())
  )
```

## Duplicate analytical keys

Duplicates are evaluated after applying the date window and removing rows with missing key components. “Rows involved” counts every row belonging to a duplicated key; these are the rows excluded by the analysis notebooks.

```{r}
duplicate_summary <- bind_rows(
  daily_duplicate_summary, gps_5min_duplicate_summary, myair_5min_duplicate_summary,
  smartwatch_high_5min_duplicate_summary, smartwatch_low_5min_duplicate_summary
) %>% arrange(factor(resolution, levels = c("5 minutes", "Daily", "Participant-night")), data_table)

duplicate_summary %>%
  mutate(percent_rows_involved = round(percent_rows_involved, 3)) %>%
  knitr::kable(
    format = "pipe", big.mark = ",",
    col.names = c(
      "Data table", "Resolution", "Duplicated keys", "Rows involved",
      "Rows with complete keys in window", "Rows involved (%)"
    ),
    caption = "Duplicated analytical keys within the retained date window"
  )
```

## Interpretation

This report identifies where records are excluded by the shared temporal and key-uniqueness rules. It does not decide whether non-duplicated sensor values are plausible, complete enough for a specific model, or clinically meaningful; those decisions remain analysis-specific.
