---
title: "Exploratory exposure-response relationships"
author: "Alessandro Fuschi"
editor: source
---

## Setup

```{r, message=FALSE}
library(tidyverse)

cfg <- list(
  data_dir = file.path("data", "5min"),
  timezone = "UTC",
  analysis_start_date = as.Date("2025-03-01"),
  hex_bins = 120,
  particle_mass_max = 1000
)
```

## Data import

The five-minute tables are read using `userId` and `bucket_5min` as analytical keys. Only buckets whose UTC date is strictly after 1 March 2025 and strictly before the current date are retained.

Each available bucket counts once, irrespective of the number of raw measurements produced inside the five-minute interval.

Rows associated with duplicated analytical keys are removed completely.

```{r, message=FALSE}
air <- read_csv(
  file.path(cfg$data_dir, "myair_5min.csv.gz"),
  na = c("", "NA", "NULL"), locale = locale(tz = cfg$timezone),
  col_types = cols(
    userId = col_character(),
    bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")
  ),
  show_col_types = FALSE
) %>%
  filter(
    as.Date(bucket_5min, tz = cfg$timezone) > cfg$analysis_start_date,
    as.Date(bucket_5min, tz = cfg$timezone) < Sys.Date()
  ) %>%
  add_count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n == 1) %>%
  select(-key_n)

swh <- read_csv(
  file.path(cfg$data_dir, "smartwatchhigh_5min.csv.gz"),
  na = c("", "NA", "NULL"), locale = locale(tz = cfg$timezone),
  col_types = cols(
    userId = col_character(),
    bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")
  ),
  show_col_types = FALSE
) %>%
  filter(
    as.Date(bucket_5min, tz = cfg$timezone) > cfg$analysis_start_date,
    as.Date(bucket_5min, tz = cfg$timezone) < Sys.Date()
  ) %>%
  add_count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n == 1) %>%
  select(-key_n)

swl <- read_csv(
  file.path(cfg$data_dir, "smartwatchlow_5min.csv.gz"),
  na = c("", "NA", "NULL"), locale = locale(tz = cfg$timezone),
  col_types = cols(
    userId = col_character(),
    bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")
  ),
  show_col_types = FALSE
) %>%
  filter(
    as.Date(bucket_5min, tz = cfg$timezone) > cfg$analysis_start_date,
    as.Date(bucket_5min, tz = cfg$timezone) < Sys.Date()
  ) %>%
  add_count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n == 1) %>%
  select(-key_n)
```

## Dataset construction

Environmental, physiological and activity measurements are joined using `userId` and the five-minute time bucket.

For steps, `step_mean` is used instead of `step_sum`, because the same step value is repeated in the raw observations within each bucket.

```{r}
dat <- full_join(
  air %>%
    transmute(
      userId,
      bucket_5min,
      temperature = temperature_mean,
      temperature_min,
      temperature_max,
      humidity = humidity_mean,
      humidity_min,
      humidity_max,
      pm1 = pm1_mean,
      pm25 = pm25_mean,
      pm10 = pm10_mean,
      pc03 = pc03_mean,
      pc05 = pc05_mean,
      pc1 = pc1_mean,
      pc25 = pc25_mean,
      pc5 = pc5_mean,
      pc10 = pc10_mean,
      uvb = uvb_mean,
      light = light_mean
    ),
  swh %>%
    transmute(
      userId,
      bucket_5min,
      heartrate = heartrate_mean,
      heartrate_min,
      heartrate_max,
      breathrate = breathrate_mean,
      breathrate_min,
      breathrate_max,
      oxygen = oxygens_mean,
      oxygen_min = oxygens_min,
      oxygen_max = oxygens_max
    ),
  by = c("userId", "bucket_5min"),
  relationship = "one-to-one"
) %>%
  full_join(
    swl %>%
      transmute(
        userId,
        bucket_5min,
        steps = step_mean,
        bphigh = bphigh_mean,
        bplow = bplow_mean,
        bodytemp = bodytemp_mean,
        bodytemp_min,
        bodytemp_max,
        skintemp = skintemp_mean,
        skintemp_min,
        skintemp_max
      ),
    by = c("userId", "bucket_5min"),
    relationship = "one-to-one"
  ) %>%
  arrange(userId, bucket_5min) %>%
  mutate(userId = factor(userId))
```

## Particle quality control

Particle counts are cumulative above increasing size thresholds and should be
monotonically decreasing. The same exploratory particulate-mass threshold used
in `particle_quality_control.qmd` is applied here. When either check fails,
the complete particle block is set to missing.

```{r}
dat <- dat %>%
  mutate(
    invalid_particle_order = coalesce(
      pc03 < pc05 |
        pc05 < pc1 |
        pc1 < pc25 |
        pc25 < pc5 |
        pc5 < pc10,
      FALSE
    ),

    invalid_particle_mass = coalesce(
      pm1 > cfg$particle_mass_max |
        pm25 > cfg$particle_mass_max |
        pm10 > cfg$particle_mass_max,
      FALSE
    ),

    invalid_particles =
      invalid_particle_order |
      invalid_particle_mass,

    pm1 = if_else(invalid_particles, NA_real_, pm1),
    pm25 = if_else(invalid_particles, NA_real_, pm25),
    pm10 = if_else(invalid_particles, NA_real_, pm10),

    pc03 = if_else(invalid_particles, NA_real_, pc03),
    pc05 = if_else(invalid_particles, NA_real_, pc05),
    pc1 = if_else(invalid_particles, NA_real_, pc1),
    pc25 = if_else(invalid_particles, NA_real_, pc25),
    pc5 = if_else(invalid_particles, NA_real_, pc5),
    pc10 = if_else(invalid_particles, NA_real_, pc10)
  )
```

## Derived variables

```{r}
dat <- dat %>%
  mutate(
    humidex = ThermIndex::humidex(
      temp = temperature,
      ur = humidity
    ),
    heartrate_range = heartrate_max - heartrate_min,
    breathrate_range = breathrate_max - breathrate_min,
    oxygen_range = oxygen_max - oxygen_min,
    bodytemp_range = bodytemp_max - bodytemp_min,
    skintemp_range = skintemp_max - skintemp_min,
    pulse_pressure = bphigh - bplow,
    mean_arterial_pressure = (bphigh + 2 * bplow) / 3
  )
```

## Within-subject centering

Physiological variables are expressed as deviations from each subject's personal mean.

```{r}
dat <- dat %>%
  group_by(userId) %>%
  mutate(
    heartrate_c =
      heartrate - mean(heartrate, na.rm = TRUE),

    heartrate_range_c =
      heartrate_range - mean(heartrate_range, na.rm = TRUE),

    breathrate_c =
      breathrate - mean(breathrate, na.rm = TRUE),

    oxygen_min_c =
      oxygen_min - mean(oxygen_min, na.rm = TRUE),

    skintemp_c =
      skintemp - mean(skintemp, na.rm = TRUE),

    pulse_pressure_c =
      pulse_pressure - mean(pulse_pressure, na.rm = TRUE),

    mean_arterial_pressure_c =
      mean_arterial_pressure -
      mean(mean_arterial_pressure, na.rm = TRUE)
  ) %>%
  ungroup()
```

## Exposure-response hexplots

The plots involving particulate matter use the exploratory quality-control rule
defined above. The threshold should be validated against the sensor
specification before formal inference. Particulate matter, UVB, and light use a
`log1p` horizontal scale; all other axes retain their original units.

### Steps and heart rate

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(steps), !is.na(heartrate_c)) %>%
  ggplot(aes(x = steps, y = heartrate_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Steps and heart rate",
    subtitle = "Five-minute observations",
    x = "Steps in five minutes",
    y = "Heart rate relative to personal mean (bpm)"
  ) +
  theme_minimal()
```

### Temperature and heart rate

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(temperature), !is.na(heartrate_c)) %>%
  ggplot(aes(x = temperature, y = heartrate_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Temperature and heart rate",
    subtitle = "Five-minute observations",
    x = "Temperature (°C)",
    y = "Heart rate relative to personal mean (bpm)"
  ) +
  theme_minimal()
```

### Humidex and heart rate

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(humidex), !is.na(heartrate_c)) %>%
  ggplot(aes(x = humidex, y = heartrate_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Humidex and heart rate",
    subtitle = "Five-minute observations",
    x = "Humidex",
    y = "Heart rate relative to personal mean (bpm)"
  ) +
  theme_minimal()
```

### Humidex and heart-rate range

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(humidex), !is.na(heartrate_range_c)) %>%
  ggplot(aes(x = humidex, y = heartrate_range_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Humidex and heart-rate range",
    subtitle = "Five-minute observations",
    x = "Humidex",
    y = "Heart-rate range relative to personal mean (bpm)"
  ) +
  theme_minimal()
```

### Temperature and skin temperature

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(temperature), !is.na(skintemp_c)) %>%
  ggplot(aes(x = temperature, y = skintemp_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Temperature and skin temperature",
    subtitle = "Five-minute observations",
    x = "Temperature (°C)",
    y = "Skin temperature relative to personal mean (°C)"
  ) +
  theme_minimal()
```

### Humidex and skin temperature

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(humidex), !is.na(skintemp_c)) %>%
  ggplot(aes(x = humidex, y = skintemp_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Humidex and skin temperature",
    subtitle = "Five-minute observations",
    x = "Humidex",
    y = "Skin temperature relative to personal mean (°C)"
  ) +
  theme_minimal()
```

### Humidex and breathing rate

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(humidex), !is.na(breathrate_c)) %>%
  ggplot(aes(x = humidex, y = breathrate_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Humidex and breathing rate",
    subtitle = "Five-minute observations",
    x = "Humidex",
    y = "Breathing rate relative to personal mean"
  ) +
  theme_minimal()
```

### Humidex and pulse pressure

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(humidex), !is.na(pulse_pressure_c)) %>%
  ggplot(aes(x = humidex, y = pulse_pressure_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Humidex and pulse pressure",
    subtitle = "Five-minute observations",
    x = "Humidex",
    y = "Pulse pressure relative to personal mean (mmHg)"
  ) +
  theme_minimal()
```

### Humidex and mean arterial pressure

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(humidex), !is.na(mean_arterial_pressure_c)) %>%
  ggplot(aes(x = humidex, y = mean_arterial_pressure_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Humidex and mean arterial pressure",
    subtitle = "Five-minute observations",
    x = "Humidex",
    y = "Mean arterial pressure relative to personal mean (mmHg)"
  ) +
  theme_minimal()
```

### PM2.5 and breathing rate

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(pm25), !is.na(breathrate_c)) %>%
  ggplot(aes(x = pm25, y = breathrate_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_continuous(transform = "log1p") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "PM2.5 and breathing rate",
    subtitle = "Five-minute observations",
    x = "PM2.5 (µg/m³)",
    y = "Breathing rate relative to personal mean"
  ) +
  theme_minimal()
```

### PM2.5 and oxygen saturation

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(pm25), !is.na(oxygen_min_c)) %>%
  ggplot(aes(x = pm25, y = oxygen_min_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_continuous(transform = "log1p") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "PM2.5 and minimum oxygen saturation",
    subtitle = "Five-minute observations",
    x = "PM2.5 (µg/m³)",
    y = "Minimum oxygen saturation relative to personal mean"
  ) +
  theme_minimal()
```

### PM2.5 and heart rate

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(pm25), !is.na(heartrate_c)) %>%
  ggplot(aes(x = pm25, y = heartrate_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_continuous(transform = "log1p") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "PM2.5 and heart rate",
    subtitle = "Five-minute observations",
    x = "PM2.5 (µg/m³)",
    y = "Heart rate relative to personal mean (bpm)"
  ) +
  theme_minimal()
```

### UVB and heart rate

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(uvb), !is.na(heartrate_c)) %>%
  ggplot(aes(x = uvb, y = heartrate_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_continuous(transform = "log1p") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "UVB and heart rate",
    subtitle = "Five-minute observations",
    x = "UVB",
    y = "Heart rate relative to personal mean (bpm)"
  ) +
  theme_minimal()
```

### Light and skin temperature

```{r, fig.width=10, fig.height=6.5}
dat %>%
  filter(!is.na(light), !is.na(skintemp_c)) %>%
  ggplot(aes(x = light, y = skintemp_c)) +
  geom_hex(bins = cfg$hex_bins) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_x_continuous(transform = "log1p") +
  scale_fill_distiller(
    palette = "Spectral", direction = -1, transform = "log10",
    breaks = c(1, 10, 100, 1000, 10000, 100000),
    labels = scales::label_number(big.mark = ".", decimal.mark = ","),
    name = "Observations\n(log10 scale)"
  ) +
  labs(
    title = "Light and skin temperature",
    subtitle = "Five-minute observations",
    x = "Light",
    y = "Skin temperature relative to personal mean (°C)"
  ) +
  theme_minimal()
```
