---
title: "Five-minute bucket overlap"
author: "Alessandro Fuschi"
editor: source
toc: true
---

## Setup

```{r, message=FALSE}
library(tidyverse)
library(ComplexUpset)

cfg <- list(
  data_dir = file.path("data", "5min"),
  analysis_start_date = as.Date("2025-03-01"),
  timezone = "UTC",
  upset_top_intersections = 20
)

theme_set(theme_minimal(base_size = 12))
```

## Aim

This report describes how often the four high-resolution data streams overlap. Two observations overlap only when they belong to the same participant and the same UTC five-minute bucket. The counts therefore represent matched participant-buckets rather than pairs of raw sensor readings.

The first view shows all pairwise intersections. The second partitions the union of the data into exact combinations of available streams, providing the information of an UpSet plot in a compact form suitable for four sets.

The report then repeats the comparison using the availability of 19 individual variables. Their origin is retained explicitly, and the variable-level UpSet plot shows the 20 most frequent exact intersections. This limit affects only presentation: all intersections are calculated before ranking.

This is a coverage analysis: an overlap identifies buckets that could contribute to a paired analysis, but it does not verify whether particular measurements inside those buckets are non-missing or valid.

## Data import

Only the analytical keys and variables used in the overlap analysis are imported. The shared project inclusion rules are applied without repeating their diagnostics, which are available in the [data-quality report](data_quality_control.html).

```{r, message=FALSE}
gps <- read_csv(
  file.path(cfg$data_dir, "gps_5min.csv.gz"),
  col_select = c(userId, bucket_5min, longitude_mean, latitude_mean),
  col_types = cols(userId = col_character(), bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")),
  locale = locale(tz = cfg$timezone), na = c("", "NA", "NULL"), show_col_types = FALSE
) %>%
  filter(
    !is.na(userId), !is.na(bucket_5min),
    as.Date(bucket_5min, tz = cfg$timezone) > cfg$analysis_start_date,
    as.Date(bucket_5min, tz = cfg$timezone) < Sys.Date()
  ) %>%
  add_count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n == 1) %>%
  select(-key_n)

air <- read_csv(
  file.path(cfg$data_dir, "myair_5min.csv.gz"),
  col_select = c(
    userId, bucket_5min, pm1_mean, pm25_mean, pm10_mean, temperature_mean,
    humidity_mean, pressure_mean, sound_mean, uvb_mean, light_mean
  ),
  col_types = cols(userId = col_character(), bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")),
  locale = locale(tz = cfg$timezone), na = c("", "NA", "NULL"), show_col_types = FALSE
) %>%
  filter(
    !is.na(userId), !is.na(bucket_5min),
    as.Date(bucket_5min, tz = cfg$timezone) > cfg$analysis_start_date,
    as.Date(bucket_5min, tz = cfg$timezone) < Sys.Date()
  ) %>%
  add_count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n == 1) %>%
  select(-key_n)

smartwatch_high <- read_csv(
  file.path(cfg$data_dir, "smartwatchhigh_5min.csv.gz"),
  col_select = c(userId, bucket_5min, heartrate_mean, oxygens_mean, breathrate_mean),
  col_types = cols(userId = col_character(), bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")),
  locale = locale(tz = cfg$timezone), na = c("", "NA", "NULL"), show_col_types = FALSE
) %>%
  filter(
    !is.na(userId), !is.na(bucket_5min),
    as.Date(bucket_5min, tz = cfg$timezone) > cfg$analysis_start_date,
    as.Date(bucket_5min, tz = cfg$timezone) < Sys.Date()
  ) %>%
  add_count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n == 1) %>%
  select(-key_n)

smartwatch_low <- read_csv(
  file.path(cfg$data_dir, "smartwatchlow_5min.csv.gz"),
  col_select = c(
    userId, bucket_5min, step_mean, cal_mean, bphigh_mean, bplow_mean,
    bodytemp_mean, skintemp_mean
  ),
  col_types = cols(userId = col_character(), bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")),
  locale = locale(tz = cfg$timezone), na = c("", "NA", "NULL"), show_col_types = FALSE
) %>%
  filter(
    !is.na(userId), !is.na(bucket_5min),
    as.Date(bucket_5min, tz = cfg$timezone) > cfg$analysis_start_date,
    as.Date(bucket_5min, tz = cfg$timezone) < Sys.Date()
  ) %>%
  add_count(userId, bucket_5min, name = "key_n") %>%
  filter(key_n == 1) %>%
  select(-key_n)

if (anyDuplicated(gps) > 0 || anyDuplicated(air) > 0 ||
    anyDuplicated(smartwatch_high) > 0 || anyDuplicated(smartwatch_low) > 0) {
  stop("At least one data stream contains duplicated analytical keys after filtering.")
}
```

## Overlap dataset

The four sets are joined without discarding buckets that occur in only one stream. Missing membership flags are replaced by `FALSE`, so every row describes the complete availability pattern for one participant-bucket.

```{r}
overlap <- gps %>%
  transmute(
    userId, bucket_5min, gps = TRUE,
    gps_position = !is.na(longitude_mean) & !is.na(latitude_mean)
  ) %>%
  full_join(
    air %>%
      transmute(
        userId, bucket_5min, environment = TRUE,
        pm1 = !is.na(pm1_mean), pm25 = !is.na(pm25_mean), pm10 = !is.na(pm10_mean),
        air_temperature = !is.na(temperature_mean), humidity = !is.na(humidity_mean),
        pressure = !is.na(pressure_mean), sound = !is.na(sound_mean),
        uvb = !is.na(uvb_mean), light = !is.na(light_mean)
      ),
    by = c("userId", "bucket_5min"), relationship = "one-to-one"
  ) %>%
  full_join(
    smartwatch_high %>%
      transmute(
        userId, bucket_5min, smartwatch_high = TRUE,
        heart_rate = !is.na(heartrate_mean), oxygen_saturation = !is.na(oxygens_mean),
        breathing_rate = !is.na(breathrate_mean)
      ),
    by = c("userId", "bucket_5min"), relationship = "one-to-one"
  ) %>%
  full_join(
    smartwatch_low %>%
      transmute(
        userId, bucket_5min, smartwatch_low = TRUE,
        steps = !is.na(step_mean), calories = !is.na(cal_mean),
        systolic_blood_pressure = !is.na(bphigh_mean),
        diastolic_blood_pressure = !is.na(bplow_mean),
        body_temperature = !is.na(bodytemp_mean), skin_temperature = !is.na(skintemp_mean)
      ),
    by = c("userId", "bucket_5min"), relationship = "one-to-one"
  ) %>%
  mutate(
    gps = replace_na(gps, FALSE), environment = replace_na(environment, FALSE),
    smartwatch_high = replace_na(smartwatch_high, FALSE),
    smartwatch_low = replace_na(smartwatch_low, FALSE),
    across(gps_position:skin_temperature, ~ replace_na(.x, FALSE))
  )

rm(gps, air, smartwatch_high, smartwatch_low)

if (anyDuplicated(overlap[c("userId", "bucket_5min")]) > 0) {
  stop("The combined overlap dataset contains duplicated participant-bucket keys.")
}
```

## Stream totals

The total for each stream is the denominator used to interpret the pairwise intersections. A participant contributes at most once per stream and five-minute bucket.

```{r}
stream_summary <- tibble(
  data_stream = c("GPS", "Environment", "Smartwatch high frequency", "Smartwatch low frequency"),
  observed_5min_buckets = c(
    sum(overlap$gps), sum(overlap$environment),
    sum(overlap$smartwatch_high), sum(overlap$smartwatch_low)
  ),
  participants = c(
    n_distinct(overlap$userId[overlap$gps]),
    n_distinct(overlap$userId[overlap$environment]),
    n_distinct(overlap$userId[overlap$smartwatch_high]),
    n_distinct(overlap$userId[overlap$smartwatch_low])
  )
)

stream_summary %>%
  knitr::kable(
    format = "pipe", big.mark = ",",
    col.names = c("Data stream", "Observed 5-minute buckets", "Participants"),
    caption = "Five-minute bucket totals by data stream"
  )
```

## Pairwise intersections

The diagonal contains the total number of buckets in each stream. Every off-diagonal cell contains the number of participant-buckets shared by the corresponding pair. A bucket available in three or four streams contributes to multiple pairwise cells.

```{r}
stream_levels <- c("GPS", "Environment", "Smartwatch high frequency", "Smartwatch low frequency")

pairwise_matrix <- overlap %>%
  select(gps, environment, smartwatch_high, smartwatch_low) %>%
  set_names(stream_levels) %>%
  as.matrix() %>%
  crossprod()

pairwise_overlap <- as.data.frame(as.table(pairwise_matrix), stringsAsFactors = FALSE) %>%
  as_tibble() %>%
  rename(stream_x = Var1, stream_y = Var2, overlapping_buckets = Freq) %>%
  mutate(
    stream_x = factor(stream_x, levels = stream_levels),
    stream_y = factor(stream_y, levels = rev(stream_levels)),
    light_label = overlapping_buckets > 0.45 * max(overlapping_buckets)
  )
```

```{r, fig.width=10, fig.height=8}
pairwise_overlap %>%
  ggplot(aes(x = stream_x, y = stream_y, fill = overlapping_buckets)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(
    aes(label = scales::label_number(big.mark = ",", accuracy = 1)(overlapping_buckets), color = light_label),
    fontface = "bold", size = 4
  ) +
  scale_fill_viridis_c(
    option = "C", direction = -1, transform = "sqrt",
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  coord_equal() +
  labs(
    title = "Pairwise overlap of five-minute buckets",
    subtitle = "Same participant and UTC bucket; diagonal cells are stream totals",
    x = NULL, y = NULL, fill = "Overlapping\nbuckets"
  ) +
  theme(
    panel.grid = element_blank(),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

### Unique stream pairs

The percentage uses the smaller stream as denominator. It answers what proportion of the maximum possible pairwise overlap was observed, while the bucket count remains the primary result.

```{r}
pairwise_table <- pairwise_overlap %>%
  mutate(
    stream_x_index = match(as.character(stream_x), stream_levels),
    stream_y_index = match(as.character(stream_y), stream_levels)
  ) %>%
  filter(stream_x_index < stream_y_index) %>%
  left_join(
    stream_summary %>% select(stream_x = data_stream, stream_x_total = observed_5min_buckets),
    by = c("stream_x" = "stream_x")
  ) %>%
  left_join(
    stream_summary %>% select(stream_y = data_stream, stream_y_total = observed_5min_buckets),
    by = c("stream_y" = "stream_y")
  ) %>%
  transmute(
    data_stream_pair = paste(stream_x, stream_y, sep = " — "),
    overlapping_buckets,
    percent_of_smaller_stream = 100 * overlapping_buckets / pmin(stream_x_total, stream_y_total)
  ) %>%
  arrange(desc(overlapping_buckets))

pairwise_table %>%
  mutate(percent_of_smaller_stream = round(percent_of_smaller_stream, 1)) %>%
  knitr::kable(
    format = "pipe", big.mark = ",",
    col.names = c("Data stream pair", "Overlapping 5-minute buckets", "Smaller stream covered (%)"),
    caption = "Pairwise five-minute bucket intersections"
  )
```

## Exact intersection patterns

Unlike the pairwise matrix, the exact combinations partition the union: each participant-bucket appears in one bar only. For example, a bucket labelled “Environment + Smartwatch high frequency” has those two streams and neither GPS nor smartwatch low-frequency data.

```{r}
intersection_summary <- overlap %>%
  count(gps, environment, smartwatch_high, smartwatch_low, name = "buckets") %>%
  filter(buckets > 0) %>%
  rowwise() %>%
  mutate(
    data_streams = paste(
      stream_levels[c_across(c(gps, environment, smartwatch_high, smartwatch_low))],
      collapse = " + "
    ),
    streams_in_combination = sum(c_across(c(gps, environment, smartwatch_high, smartwatch_low))),
    percent_of_union = 100 * buckets / nrow(overlap)
  ) %>%
  ungroup() %>%
  arrange(buckets)
```

```{r, fig.width=12, fig.height=9}
intersection_summary %>%
  mutate(data_streams = fct_reorder(data_streams, buckets)) %>%
  ggplot(aes(x = buckets, y = data_streams, fill = factor(streams_in_combination))) +
  geom_col() +
  geom_text(
    aes(label = paste0(scales::label_number(big.mark = ",", accuracy = 1)(buckets), " (", round(percent_of_union, 1), "%)")),
    hjust = -0.05, size = 3.5
  ) +
  scale_x_continuous(
    labels = scales::label_number(big.mark = ","),
    expand = expansion(mult = c(0, 0.22))
  ) +
  scale_fill_viridis_d(option = "C", direction = -1) +
  labs(
    title = "Exact five-minute intersection patterns",
    subtitle = "Each participant-bucket belongs to exactly one combination",
    x = "Participant-buckets", y = NULL, fill = "Streams in\ncombination"
  ) +
  theme(panel.grid.major.y = element_blank())
```

## Variable-level availability

A variable is available when its value is non-missing in a participant-bucket. GPS position requires both longitude and latitude. The selected variables cover location, the main environmental measurements and the physiological or activity measurements currently used in the exploratory analyses.

```{r}
variable_metadata <- tribble(
  ~variable, ~variable_label, ~origin,
  "gps_position", "Position", "GPS",
  "pm1", "PM1", "Environment",
  "pm25", "PM2.5", "Environment",
  "pm10", "PM10", "Environment",
  "air_temperature", "Air temperature", "Environment",
  "humidity", "Humidity", "Environment",
  "pressure", "Pressure", "Environment",
  "sound", "Sound", "Environment",
  "uvb", "UVB", "Environment",
  "light", "Light", "Environment",
  "heart_rate", "Heart rate", "Smartwatch high frequency",
  "oxygen_saturation", "Oxygen saturation", "Smartwatch high frequency",
  "breathing_rate", "Breathing rate", "Smartwatch high frequency",
  "steps", "Steps", "Smartwatch low frequency",
  "calories", "Calories", "Smartwatch low frequency",
  "systolic_blood_pressure", "Systolic blood pressure", "Smartwatch low frequency",
  "diastolic_blood_pressure", "Diastolic blood pressure", "Smartwatch low frequency",
  "body_temperature", "Body temperature", "Smartwatch low frequency",
  "skin_temperature", "Skin temperature", "Smartwatch low frequency"
) %>%
  mutate(
    origin = factor(origin, levels = stream_levels),
    origin_code = recode(
      as.character(origin), GPS = "GPS", Environment = "ENV",
      `Smartwatch high frequency` = "SWH", `Smartwatch low frequency` = "SWL"
    ),
    upset_label = paste(origin_code, variable_label, sep = " · ")
  )

variable_names <- variable_metadata$variable
variable_matrix <- overlap %>% select(all_of(variable_names)) %>% as.matrix()
storage.mode(variable_matrix) <- "double"

variable_summary <- variable_metadata %>%
  mutate(observed_5min_buckets = as.numeric(colSums(variable_matrix))) %>%
  select(origin, variable = variable_label, observed_5min_buckets)

variable_summary %>%
  knitr::kable(
    format = "pipe", big.mark = ",",
    col.names = c("Origin", "Variable", "Observed 5-minute buckets"),
    caption = "Five-minute bucket totals by variable"
  )
```

## Pairwise variable co-occurrence

The variable-level tile counts participant-buckets in which both values are available. Facet strips identify the origin of the variables and separate the matrix into source blocks. Labels use compact `K` and `M` notation to remain readable; the colour scale and underlying counts retain the absolute bucket numbers.

```{r}
pairwise_variable_matrix <- crossprod(variable_matrix)
dimnames(pairwise_variable_matrix) <- list(variable_names, variable_names)

pairwise_variable_overlap <- as.data.frame(as.table(pairwise_variable_matrix), stringsAsFactors = FALSE) %>%
  as_tibble() %>%
  rename(variable_x = Var1, variable_y = Var2, overlapping_buckets = Freq) %>%
  left_join(
    variable_metadata %>%
      select(variable_x = variable, variable_x_label = variable_label, origin_x = origin),
    by = "variable_x", relationship = "many-to-one"
  ) %>%
  left_join(
    variable_metadata %>%
      select(variable_y = variable, variable_y_label = variable_label, origin_y = origin),
    by = "variable_y", relationship = "many-to-one"
  ) %>%
  mutate(
    variable_x_label = factor(variable_x_label, levels = variable_metadata$variable_label),
    variable_y_label = factor(variable_y_label, levels = rev(variable_metadata$variable_label)),
    origin_x = factor(
      origin_x, levels = stream_levels,
      labels = c("GPS", "Environment", "Smartwatch HF", "Smartwatch LF")
    ),
    origin_y = factor(
      origin_y, levels = rev(stream_levels),
      labels = rev(c("GPS", "Environment", "Smartwatch HF", "Smartwatch LF"))
    ),
    light_label = overlapping_buckets > 0.45 * max(overlapping_buckets)
  )
```

```{r, fig.width=15, fig.height=13}
pairwise_variable_overlap %>%
  ggplot(aes(x = variable_x_label, y = variable_y_label, fill = overlapping_buckets)) +
  geom_tile(color = "white", linewidth = 0.35) +
  geom_text(
    aes(
      label = scales::label_number(
        scale_cut = scales::cut_short_scale(), accuracy = 0.1
      )(overlapping_buckets),
      color = light_label
    ),
    fontface = "bold", size = 4.2
  ) +
  facet_grid(origin_y ~ origin_x, scales = "free", space = "free") +
  scale_fill_viridis_c(
    option = "C", direction = -1, transform = "sqrt",
    labels = scales::label_number(big.mark = ",")
  ) +
  scale_color_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  labs(
    title = "Pairwise co-occurrence of variables in five-minute buckets",
    subtitle = "Same participant and UTC bucket; facet strips show variable origin",
    x = NULL, y = NULL, fill = "Co-occurring\nbuckets"
  ) +
  theme(
    panel.grid = element_blank(), panel.spacing = grid::unit(0.3, "lines"),
    axis.text.x = element_text(angle = 60, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10), strip.text = element_text(size = 11, face = "bold"),
    plot.title = element_text(size = 16), plot.subtitle = element_text(size = 12),
    legend.title = element_text(size = 11), legend.text = element_text(size = 10)
  )
```

## Variable UpSet plot

The UpSet plot is produced with `ComplexUpset` and shows the 20 most frequent exclusive availability patterns among the selected variables. Before plotting, millions of participant-buckets are collapsed into their distinct Boolean patterns and the original bucket count is retained as a weight. All patterns are ranked, but ComplexUpset receives only the displayed rows, while the bars still report absolute participant-bucket counts. This avoids the package's costly expansion of manually specified intersections across 19 sets. Variable labels use the prefixes `GPS`, `ENV`, `SWH` and `SWL` to retain their source.

```{r}
pattern_code <- as.integer(drop(variable_matrix %*% bitwShiftL(1L, seq_along(variable_names) - 1L)))
pattern_counts <- tabulate(pattern_code + 1L, nbins = 2^length(variable_names))
observed_pattern_codes <- which(pattern_counts > 0L) - 1L
observed_pattern_codes <- observed_pattern_codes[observed_pattern_codes > 0L]
bit_values <- bitwShiftL(1L, seq_along(variable_names) - 1L)
pattern_matrix <- outer(observed_pattern_codes, bit_values, bitwAnd) > 0L

variable_upset_patterns <- pattern_matrix %>%
  as.data.frame() %>%
  set_names(variable_metadata$upset_label) %>%
  mutate(bucket_n = pattern_counts[observed_pattern_codes + 1L]) %>%
  arrange(desc(bucket_n))

observed_pattern_n <- nrow(variable_upset_patterns)
top_pattern_n <- min(cfg$upset_top_intersections, observed_pattern_n)
variable_upset_plot_data <- variable_upset_patterns %>% slice_head(n = top_pattern_n)
top_intersection_share <- 100 *
  sum(variable_upset_plot_data$bucket_n) / sum(variable_upset_patterns$bucket_n)

rm(variable_matrix, overlap, pattern_code, pattern_counts, pattern_matrix, variable_upset_patterns)
invisible(gc())
```

```{r, fig.width=16, fig.height=12, warning=FALSE}
upset(
  variable_upset_plot_data,
  intersect = variable_metadata$upset_label,
  intersections = "observed",
  name = "Variable intersection",
  mode = "distinct",
  sort_sets = FALSE,
  sort_intersections = FALSE,
  height_ratio = 0.55,
  base_annotations = list(),
  annotations = list(
    "Intersection size" = ggplot(mapping = aes(x = intersection, y = bucket_n)) +
      stat_summary(
        fun = sum, geom = "col",
        fill = viridisLite::viridis(1, option = "C", begin = 0.65)
      ) +
      stat_summary(
        aes(label = after_stat(scales::comma(y))),
        fun = sum, geom = "text", vjust = -0.4, size = 4
      )
  ),
  set_sizes = FALSE
) +
  patchwork::plot_annotation(
    title = "Most frequent exact variable intersections",
    subtitle = paste0(
      "Top ", top_pattern_n, " of ", observed_pattern_n, " patterns; ",
      round(top_intersection_share, 1), "% of variable-observed participant-buckets"
    )
  )
```

## Interpretation boundaries

The pairwise counts describe the available sample size for analyses that combine two data streams or variables. They do not imply a statistical association, and repeated buckets from the same participant are not independent observations. Variable availability is based on non-missing values; analysis-specific validity rules may reduce the usable sample further.
