---
title: "Coverage by participant and nationality"
author: "Alessandro Fuschi"
editor: source
toc: true
---

## Setup

```{r, message=FALSE}
library(tidyverse)

cfg <- list(
  accounts_dir = file.path("data", "accounts"),
  daily_dir = file.path("data", "daily"),
  tidy_dir = file.path("data", "tidy"),
  analysis_start_date = as.Date("2025-03-01"),
  minutes_per_bucket = 5,
  minutes_per_day = 24 * 60
)

theme_set(theme_minimal(base_size = 12))
```

## Aim and privacy

This exploratory report compares data coverage across nationalities and participants.

The primary unit is the five-minute bucket. One or multiple raw measurements from the same participant and data stream inside a bucket count as one observed participant-bucket. Raw record counts are intentionally not compared because the instruments have different sampling frequencies.

Only `userId` and nationality are imported from the account table. Email and login information are not read. Real user identifiers are replaced by deterministic labels before any participant-level result is displayed; these labels are pseudonyms for this report and should not be interpreted as full anonymisation.

## Data import

```{r, message=FALSE}
accounts <- read_csv(
  file.path(cfg$accounts_dir, "active_accounts.csv.gz"),
  col_select = c(userId, nationality),
  col_types = cols(userId = col_character(), nationality = col_character()),
  na = c("", "NA", "NULL"), show_col_types = FALSE
)

gps_raw <- read_csv(
  file.path(cfg$daily_dir, "gps_daily.csv.gz"),
  col_select = c(userId, date, five_min_n),
  col_types = cols(userId = col_character(), date = col_date(), five_min_n = col_double()),
  na = c("", "NA", "NULL"), show_col_types = FALSE
)

air_raw <- read_csv(
  file.path(cfg$daily_dir, "myair_daily.csv.gz"),
  col_select = c(userId, date, five_min_n),
  col_types = cols(userId = col_character(), date = col_date(), five_min_n = col_double()),
  na = c("", "NA", "NULL"), show_col_types = FALSE
)

smartwatch_high_raw <- read_csv(
  file.path(cfg$daily_dir, "smartwatchhigh_daily.csv.gz"),
  col_select = c(userId, date, five_min_n),
  col_types = cols(userId = col_character(), date = col_date(), five_min_n = col_double()),
  na = c("", "NA", "NULL"), show_col_types = FALSE
)

smartwatch_low_raw <- read_csv(
  file.path(cfg$daily_dir, "smartwatchlow_daily.csv.gz"),
  col_select = c(userId, date, five_min_n),
  col_types = cols(userId = col_character(), date = col_date(), five_min_n = col_double()),
  na = c("", "NA", "NULL"), show_col_types = FALSE
)

sleep_raw <- read_csv(
  file.path(cfg$tidy_dir, "sleep_tidy.csv.gz"),
  col_select = c(userId, date),
  col_types = cols(userId = col_character(), date = col_date()),
  na = c("", "NA", "NULL"), show_col_types = FALSE
)
```

## Account checks

The account table must contain one row per participant and a non-missing nationality before it can be joined to the coverage tables.

```{r}
if (anyDuplicated(accounts$userId) > 0) stop("Duplicated userId values were found in active_accounts.csv.gz.")
if (any(is.na(accounts$nationality) | accounts$nationality == "")) stop("Missing nationality values were found in active_accounts.csv.gz.")

nationality_roster <- accounts %>% count(nationality, name = "participants", sort = TRUE)

nationality_roster %>%
  knitr::kable(
    format = "pipe", big.mark = ",",
    col.names = c("Nationality", "Active participants"),
    caption = "Participants in the active-account table"
  )
```

## Analysis dataset

The shared project inclusion rules are applied before coverage is calculated. Their diagnostics are collected in the dedicated data-quality report.

```{r}
gps_raw <- gps_raw %>% filter(!is.na(date), date > cfg$analysis_start_date, date < Sys.Date())
air_raw <- air_raw %>% filter(!is.na(date), date > cfg$analysis_start_date, date < Sys.Date())
smartwatch_high_raw <- smartwatch_high_raw %>% filter(!is.na(date), date > cfg$analysis_start_date, date < Sys.Date())
smartwatch_low_raw <- smartwatch_low_raw %>% filter(!is.na(date), date > cfg$analysis_start_date, date < Sys.Date())
sleep_raw <- sleep_raw %>% filter(!is.na(date), date > cfg$analysis_start_date, date < Sys.Date())
gps <- gps_raw %>% add_count(userId, date, name = "key_n") %>% filter(key_n == 1) %>% select(-key_n)
air <- air_raw %>% add_count(userId, date, name = "key_n") %>% filter(key_n == 1) %>% select(-key_n)
smartwatch_high <- smartwatch_high_raw %>% add_count(userId, date, name = "key_n") %>% filter(key_n == 1) %>% select(-key_n)
smartwatch_low <- smartwatch_low_raw %>% add_count(userId, date, name = "key_n") %>% filter(key_n == 1) %>% select(-key_n)
sleep <- sleep_raw %>% add_count(userId, date, name = "key_n") %>% filter(key_n == 1) %>% select(-key_n)

rm(gps_raw, air_raw, smartwatch_high_raw, smartwatch_low_raw, sleep_raw)
```

## Five-minute coverage dataset

The four five-minute data streams are stacked without joining them to one another. Coverage is therefore retained even when a participant has data from only one instrument on a given day.

```{r}
five_min_daily <- bind_rows(
  gps %>% select(userId, date, five_min_buckets = five_min_n) %>% mutate(data_stream = "GPS"),
  air %>% select(userId, date, five_min_buckets = five_min_n) %>% mutate(data_stream = "Environment"),
  smartwatch_high %>% select(userId, date, five_min_buckets = five_min_n) %>% mutate(data_stream = "Smartwatch high frequency"),
  smartwatch_low %>% select(userId, date, five_min_buckets = five_min_n) %>% mutate(data_stream = "Smartwatch low frequency")
)

if (any(five_min_daily$five_min_buckets < 0, na.rm = TRUE)) stop("Negative five-minute bucket counts were found.")
if (any(five_min_daily$five_min_buckets > cfg$minutes_per_day / cfg$minutes_per_bucket, na.rm = TRUE)) stop("A participant-day exceeds 288 five-minute buckets.")

five_min_daily <- five_min_daily %>%
  left_join(accounts, by = "userId", relationship = "many-to-one")

if (any(is.na(five_min_daily$nationality))) stop("At least one participant in the coverage data is missing from the active-account table.")
```

## Participant pseudonyms

Participants are ordered within nationality by their internal identifier and assigned a report-only sequential label. The original identifier remains available only for internal joins and is never printed.

```{r}
participant_key <- accounts %>%
  arrange(nationality, userId) %>%
  mutate(participant = sprintf("Participant %03d", row_number())) %>%
  select(userId, participant)

five_min_daily <- five_min_daily %>%
  left_join(participant_key, by = "userId", relationship = "many-to-one")
```

## Coverage by nationality

Participants without data in a stream are retained with zero coverage. This makes nationality comparisons refer to the complete active-account roster rather than only to contributors.

```{r}
participant_stream_observed <- five_min_daily %>%
  filter(five_min_buckets > 0) %>%
  group_by(userId, nationality, participant, data_stream) %>%
  summarise(
    active_days = n_distinct(date),
    observed_5min_buckets = sum(five_min_buckets),
    median_buckets_per_active_day = median(five_min_buckets),
    .groups = "drop"
  )

participant_stream_coverage <- crossing(
  accounts %>% left_join(participant_key, by = "userId", relationship = "one-to-one"),
  data_stream = c("GPS", "Environment", "Smartwatch high frequency", "Smartwatch low frequency")
) %>%
  left_join(
    participant_stream_observed,
    by = c("userId", "nationality", "participant", "data_stream"),
    relationship = "one-to-one"
  ) %>%
  mutate(
    active_days = replace_na(active_days, 0L),
    observed_5min_buckets = replace_na(observed_5min_buckets, 0),
    median_buckets_per_active_day = replace_na(median_buckets_per_active_day, 0)
  )

nationality_stream_summary <- participant_stream_coverage %>%
  group_by(nationality, data_stream) %>%
  summarise(
    participants = n(),
    participants_with_data = sum(observed_5min_buckets > 0),
    participant_days = sum(active_days),
    total_5min_buckets = sum(observed_5min_buckets),
    median_buckets_per_participant = median(observed_5min_buckets),
    median_buckets_per_contributor = median(observed_5min_buckets[observed_5min_buckets > 0]),
    .groups = "drop"
  ) %>%
  arrange(data_stream, nationality)

nationality_stream_summary %>%
  mutate(
    total_5min_buckets = format(total_5min_buckets, big.mark = ",", scientific = FALSE),
    median_buckets_per_participant = format(round(median_buckets_per_participant), big.mark = ",", scientific = FALSE),
    median_buckets_per_contributor = format(round(median_buckets_per_contributor), big.mark = ",", scientific = FALSE)
  ) %>%
  knitr::kable(
    format = "pipe", big.mark = ",",
    col.names = c(
      "Nationality", "Data stream", "Participants", "With data", "Participant-days",
      "Observed 5-minute buckets", "Median per participant", "Median per contributor"
    ),
    caption = "Five-minute coverage by nationality and data stream"
  )
```

### Absolute bucket totals

Absolute totals depend on both the number of participants and their individual monitoring duration. They should therefore be read together with the participant-level distributions below.

```{r, fig.width=12, fig.height=7}
nationality_stream_summary %>%
  ggplot(aes(x = nationality, y = total_5min_buckets, fill = nationality)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~ data_stream, scales = "free_y") +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Absolute five-minute bucket coverage by nationality",
    subtitle = "Totals include all contributing participants in each nationality",
    x = "Nationality", y = "Observed 5-minute buckets"
  )
```

### Participant-level distributions

Each point represents one pseudonymised participant with data in the corresponding stream. Active days describe monitoring duration, while total buckets combine duration and within-day coverage.

```{r, fig.width=12, fig.height=8}
participant_stream_coverage %>%
  filter(observed_5min_buckets > 0) %>%
  ggplot(aes(x = active_days, y = observed_5min_buckets, color = nationality)) +
  geom_point(alpha = 0.6, size = 1.8) +
  facet_wrap(~ data_stream, scales = "free") +
  scale_x_continuous(labels = scales::label_number(big.mark = ",")) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Coverage duration and bucket volume by participant",
    subtitle = "One point per participant and data stream",
    x = "Active days", y = "Observed 5-minute buckets", color = "Nationality"
  )
```

### Participant heatmap

The heatmap includes the full active-account roster. Dark cells indicate zero or low coverage; participant labels are report-only pseudonyms.

```{r, fig.width=11, fig.height=30}
participant_order <- participant_stream_coverage %>%
  group_by(nationality, participant) %>%
  summarise(all_stream_buckets = sum(observed_5min_buckets), .groups = "drop") %>%
  arrange(nationality, all_stream_buckets) %>%
  pull(participant)

participant_stream_coverage %>%
  mutate(participant = factor(participant, levels = participant_order)) %>%
  ggplot(aes(x = data_stream, y = participant, fill = observed_5min_buckets)) +
  geom_tile(color = "white", linewidth = 0.15) +
  facet_grid(nationality ~ ., scales = "free_y", space = "free_y") +
  scale_fill_viridis_c(transform = "sqrt", labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Five-minute coverage by participant and data stream",
    subtitle = "Absolute bucket counts; square-root colour scale",
    x = NULL, y = NULL, fill = "5-minute buckets"
  ) +
  theme(
    panel.grid = element_blank(), strip.text.y = element_text(angle = 0),
    axis.text.x = element_text(angle = 25, hjust = 1), axis.text.y = element_text(size = 5)
  )
```

## Daily coverage by nationality

This view preserves the calendar date and compares the absolute number of participant-buckets contributed by each nationality. Differences in group size remain part of the absolute count.

```{r}
daily_nationality_coverage <- five_min_daily %>%
  group_by(date, nationality, data_stream) %>%
  summarise(observed_5min_buckets = sum(five_min_buckets, na.rm = TRUE), .groups = "drop")

daily_nationality_grid <- crossing(
  date = seq(cfg$analysis_start_date + 1, Sys.Date() - 1, by = "day"),
  nationality = sort(unique(accounts$nationality)),
  data_stream = c("GPS", "Environment", "Smartwatch high frequency", "Smartwatch low frequency")
) %>%
  left_join(daily_nationality_coverage, by = c("date", "nationality", "data_stream"), relationship = "one-to-one") %>%
  mutate(observed_5min_buckets = replace_na(observed_5min_buckets, 0))
```

```{r, fig.width=13, fig.height=8}
daily_nationality_grid %>%
  ggplot(aes(x = date, y = observed_5min_buckets, color = nationality)) +
  geom_line(linewidth = 0.45, alpha = 0.85) +
  facet_wrap(~ data_stream, scales = "free_y", ncol = 1) +
  scale_x_date(date_breaks = "2 months", date_labels = "%b\n%Y", expand = c(0, 0)) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Daily five-minute coverage by nationality",
    subtitle = "Absolute participant-buckets",
    x = NULL, y = "Observed 5-minute buckets", color = "Nationality"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Sleep coverage

Sleep does not use five-minute buckets and is therefore reported separately as participant-nights. Participants without a sleep record are retained with zero coverage.

```{r}
sleep_participant_coverage <- accounts %>%
  left_join(participant_key, by = "userId", relationship = "one-to-one") %>%
  left_join(
    sleep %>% count(userId, name = "participant_nights"),
    by = "userId", relationship = "one-to-one"
  ) %>%
  mutate(participant_nights = replace_na(participant_nights, 0L))

sleep_nationality_summary <- sleep_participant_coverage %>%
  group_by(nationality) %>%
  summarise(
    participants = n(),
    participants_with_sleep = sum(participant_nights > 0),
    total_participant_nights = sum(participant_nights),
    median_nights_per_participant = median(participant_nights),
    median_nights_per_contributor = median(participant_nights[participant_nights > 0]),
    .groups = "drop"
  )

sleep_nationality_summary %>%
  mutate(
    median_nights_per_participant = round(median_nights_per_participant),
    median_nights_per_contributor = round(median_nights_per_contributor)
  ) %>%
  knitr::kable(
    format = "pipe", big.mark = ",",
    col.names = c(
      "Nationality", "Participants", "With sleep", "Participant-nights",
      "Median per participant", "Median per contributor"
    ),
    caption = "Sleep coverage by nationality"
  )
```

```{r, fig.width=10, fig.height=6.5}
sleep_participant_coverage %>%
  ggplot(aes(x = nationality, y = participant_nights, color = nationality)) +
  geom_boxplot(outlier.shape = NA, show.legend = FALSE) +
  geom_jitter(width = 0.16, alpha = 0.45, size = 1.5, show.legend = FALSE) +
  scale_y_continuous(labels = scales::label_number(big.mark = ",")) +
  labs(
    title = "Sleep coverage by participant and nationality",
    subtitle = "Participants without sleep records are retained at zero",
    x = "Nationality", y = "Participant-nights"
  )
```

## Interpretation boundaries

This report describes data availability rather than measurement validity. Absolute national totals are affected by nationality-group size, recruitment timing and monitoring duration. Participant-level comparisons are descriptive and should not be interpreted as differences in adherence or data quality without additional investigation.
