---
title: "Climate-health effects in TRIGGER wearable data"
subtitle: "Exploratory within-user physiological deviations during thermal exposure"
format: html
editor: source
---

```{r, message=FALSE}
library(tidyverse)
library(geosphere)
library(weathermetrics)
library(gt)

source("../../R/load_functions.R")
triggerR <- load_trigger_functions()
rm(load_trigger_functions)
dir.create("../../outputs/climate_healt", showWarnings = F)
```

```{r}
accounts <- if (!exists("accounts")) {
  
  triggerR$active_accounts() %>%
    rename(nationality = country) %>%
    mutate(nationality = case_when(
      nationality == "IT" ~ "Italy",
      nationality == "DE" ~ "Germany",
      nationality == "CH" ~ "Switzerland",
      nationality == "GR" ~ "Greece",
      TRUE ~ nationality)) %>%
    select(userId, nationality)
  
} else accounts

trigger_hourly <- if (!exists("trigger_hourly")) {
  
  triggerR$read_trigger_tables(granularity = "hourly", cores = 5) %>%
    mutate(userId = as.factor(userId)) %>%
    filter(!(userId == 61 & date == "2025-07-30" & hour == 0)) %>%
    left_join(accounts, by = "userId") %>%
    relocate(nationality, .after = userId) %>%
    select(-deviceId, -firmware) %>%
    filter(date >= "2025-04-01", date <= "2026-05-14")
  
} else trigger_hourly

trigger_daily <- if (!exists("trigger_daily")) {
  
  triggerR$read_trigger_tables(granularity = "daily", cores = 5,
                               min_obs_per_hour = 10) %>%
    mutate(userId = as.factor(userId)) %>%
    filter(!(userId == 61 & date == "2025-07-30")) %>%
    left_join(accounts, by = "userId") %>%
    relocate(nationality, .after = userId) %>%
    select(-deviceId, -firmware) %>%
    filter(date >= "2025-04-01", date <= "2026-05-14")
  
} else trigger_daily
```

```{r}
city_coords <- tribble(
  ~nationality,   ~thermal_city, ~city_lat, ~city_lon,
  "Germany",      "Augsburg",      48.3705,   10.8978,
  "Italy",        "Bologna",       44.4949,   11.3426,
  "Switzerland",  "Geneve",        46.2044,    6.1432,
  "Greece",       "Heraklion",     35.3387,   25.1442)

gps_thermal_distance <- trigger_hourly %>%
  filter(source == "gps") %>%
  group_by(userId, nationality, date, hour, measurement) %>%
  summarise(value = median(value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = measurement, values_from = value) %>%
  left_join(city_coords, by = "nationality") %>%
  mutate(
    distance_from_thermal_city_km = geosphere::distHaversine(
      p1 = cbind(longitude, latitude),
      p2 = cbind(city_lon, city_lat)
    ) / 1000,
    
    thermal_city_presence = case_when(
      is.na(latitude) | is.na(longitude) ~ "missing_gps",
      is.na(thermal_city) ~ "missing_thermal_city",
      distance_from_thermal_city_km <= 30 ~ "yes",
      distance_from_thermal_city_km <= 75 ~ "maybe",
      TRUE ~ "no"
    ),
    
    thermal_city_presence = factor(
      thermal_city_presence,
      levels = c("yes", "maybe", "no", "missing_gps", "missing_thermal_city"))) %>%
  select(-city_lat, -city_lon, -accuracy, -latitude, -longitude)

rm(city_coords)
gps_thermal_distance
```

```{r}
wearable <- trigger_hourly %>%
  filter(source != "gps",
         type %in% c("mean", "sum"),
         n_obs_valid >= 10) %>%
  select(userId, nationality, date, hour, measurement, value) %>%
  pivot_wider(names_from = "measurement",
              values_from = "value",
              values_fill = NA) %>%
  left_join(gps_thermal_distance,
            by = join_by(userId, nationality, date, hour)) %>%
  relocate(thermal_city, distance_from_thermal_city_km, thermal_city_presence,
           .after = nationality) 
```

```{r, message=FALSE}
utci <- read_csv(
  "../../data/environmental_indicators/processed/thermal_hourly_v2.csv") %>%
  select(city, date, hour, utci, utci_c, utci_stress_class) %>%
  mutate(nationality = case_when(
    city == "Augsburg" ~ "Germany",
    city == "Bologna" ~ "Italy",
    city == "Geneve" ~ "Switzerland",
    city == "Heraklion" ~ "Greece",
    TRUE ~ NA_character_))

wearable <- wearable %>%
  left_join(utci, by = join_by(nationality, date, hour)) %>%
  relocate(city:utci_stress_class, .after = nationality)

rm(utci, gps_thermal_distance)
```

```{r}
wearable <- wearable %>%
  mutate(heat_index = weathermetrics::heat.index(
    t = temperature, rh = humidity,
    temperature.metric = "celsius", output.metric = "celsius"))
```

```{r}
wearable %>%
  summarise(across(everything(), ~ mean(is.na(.x)) * 100)) %>%
  pivot_longer(everything(), names_to = "variable",
               values_to = "pct_missing") %>%
  arrange(desc(pct_missing)) %>%
  mutate(pct_missing = round(pct_missing, 1)) %>%
  gt()

```

```{r}
library(tidyverse)     # core: dplyr, ggplot2, tidyr, readr, forcats, lubridate
library(janitor)       # clean_names(), tabyl()
library(skimr)         # skim() — fast descriptive statistics
library(naniar)        # missing value visualisation
library(ggridges)      # ridge / joy plots
library(patchwork)     # multi-panel composition
library(lme4)          # mixed models (lmer)
library(lmerTest)      # p-values for lmer (Satterthwaite)
library(broom.mixed)   # tidy() for lmer objects
library(ggeffects)     # marginal effects from lmer
library(correlation)   # within-group correlations (easystats)
library(scales)        # ggplot2 axis formatting
```

```{r}
df <- wearable %>%
  clean_names() |>
  mutate(
    date   = ymd(date),
    month  = floor_date(date, "month"),
    season = case_when(
      month(date) %in% c(12, 1, 2)  ~ "Winter",
      month(date) %in% c(3, 4, 5)   ~ "Spring",
      month(date) %in% c(6, 7, 8)   ~ "Summer",
      month(date) %in% c(9, 10, 11) ~ "Autumn"
    ) |> factor(levels = c("Winter", "Spring", "Summer", "Autumn")),
    # UTCI stress class as ordered factor
    utci_stress_class = factor(utci_stress_class, levels = c(
      "Very strong cold stress", "Strong cold stress",
      "Moderate cold stress",   "Slight cold stress",
      "No thermal stress",
      "Moderate heat stress",   "Strong heat stress",
      "Very strong heat stress"
    )),
    # short labels for plots
    utci_class_short = fct_recode(utci_stress_class,
      "V.strong cold" = "Very strong cold stress",
      "Strong cold"   = "Strong cold stress",
      "Moderate cold" = "Moderate cold stress",
      "Slight cold"   = "Slight cold stress",
      "No stress"     = "No thermal stress",
      "Moderate heat" = "Moderate heat stress",
      "Strong heat"   = "Strong heat stress",
      "V.strong heat" = "Very strong heat stress"
    ),
    # hour as circular variable
    hour_sin = sin(2 * pi * hour / 24),
    hour_cos = cos(2 * pi * hour / 24),
    user_id  = factor(user_id)
  )

glimpse(df)
```

## 2. Descriptive statistics

### 2.1 Quick overview with `skimr`

```{r skim}
df |>
  select(utci_c, heartrate, breathrate, bphigh, bplow,
         bodytemp, skintemp, oxygens, steps, cal) |>
  skim()
```

### 2.2 Missing values

```{r missing-summary}
df |>
  select(
    heartrate, oxygens, breathrate, sleeprate, bphigh, bplow,
    bodytemp, skintemp, steps, cal,
    temperature, humidity, pm25, sound
  ) |>
  miss_var_summary() |>
  mutate(
    pct_miss = as.numeric(pct_miss),
    variable = fct_reorder(variable, pct_miss),
    label = paste0(round(pct_miss, 1), "%")
  ) |>
  ggplot(aes(pct_miss, variable)) +
  geom_col(aes(fill = pct_miss), show.legend = FALSE) +
  geom_text(
    aes(label = label),
    hjust = -0.1,
    size = 3,
    colour = "grey40"
  ) +
  scale_fill_gradient(low = "#93c5fd", high = "#dc2626") +
  scale_x_continuous(
    labels = label_percent(scale = 1),
    expand = expansion(mult = c(0, 0.08))
  ) +
  coord_cartesian(xlim = c(0, 100), clip = "off") +
  labs(
    title = "Percentage of missing values by variable",
    x = NULL,
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.major.y = element_blank(),
    plot.margin = margin(5.5, 25, 5.5, 5.5)
  )
```

```{r missing-upset, fig.height=5}
# Co-occurrence pattern of missing values
df |>
  select(heartrate, breathrate, bphigh, bplow,
         bodytemp, skintemp, oxygens, steps) |>
  gg_miss_upset(nsets = 8)
```

### 2.3 UTCI distribution by city

```{r utci-city}
city_palette <- c(
  "Augsburg"  = "#3b82f6",
  "Bologna"   = "#f97316",
  "Geneve"    = "#10b981",
  "Heraklion" = "#ef4444"
)

ggplot(df, aes(x = utci_c, y = city, fill = city)) +
  geom_density_ridges(alpha = 0.75, scale = 1.2, colour = "white") +
  scale_fill_manual(values = city_palette, guide = "none") +
  labs(
    title    = "UTCI distribution by city",
    subtitle = "Heraklion warmest, Geneva coldest",
    x = "UTCI (°C)", y = NULL
  ) +
  theme_minimal(base_size = 12)
```

### 2.4 Thermal stress classes — frequencies

```{r utci-classes, fig.width=7, fig.height=5}
p_hist_utci <- df |>
  count(utci_class_short) |>
  mutate(pct = n / sum(n)) |>
  ggplot(aes(utci_class_short, pct)) +
  geom_col(aes(fill = as.integer(utci_class_short)), show.legend = FALSE) +
  geom_text(aes(label = paste0(round(pct * 100, 2), "%")),
            vjust = -0.4, size = 3) +
  scale_fill_gradient2(low = "#1d4ed8", mid = "#22c55e", high = "#dc2626",
                       midpoint = 4.5) +
  scale_y_continuous(labels = label_percent()) +
  labs(title = "Distribution of UTCI thermal stress classes",
       x = NULL, y = "% observations") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 35, hjust = 1))
p_hist_utci
```

```{r}
ggsave("../../outputs/climate_healt/utci_classes_distribution.png", p_hist_utci,
       width = 7, height = 5, dpi = 300)
```


---

## 3. Heart rate × UTCI

### 3.1 Mean by class

```{r hr-by-class}
# Colour palette: cold blues → neutral green → heat reds (matching report)
bar_colours <- c(
  "V.strong cold" = "#1d4ed8",
  "Strong cold"   = "#3b82f6",
  "Moderate cold" = "#93c5fd",
  "Slight cold"   = "#bfdbfe",
  "No stress"     = "#fde68a",
  "Moderate heat" = "#fbbf24",
  "Strong heat"   = "#f97316",
  "V.strong heat" = "#ef4444"
)

hr_class <- df |>
  group_by(utci_class_short) |>
  summarise(
    mean_hr = mean(heartrate, na.rm = TRUE),
    sd_hr   = sd(heartrate,   na.rm = TRUE),
    n       = sum(!is.na(heartrate)),
    se      = sd_hr / sqrt(n),
    .groups = "drop"
  )

phr <- ggplot(hr_class, aes(utci_class_short, mean_hr, fill = utci_class_short)) +
  geom_col(width = 0.72, show.legend = FALSE) +
  geom_errorbar(
    aes(ymin = mean_hr - se, ymax = mean_hr + se),
    width = 0.22,
    colour = "grey40",
    linewidth = 0.5
  ) +
  geom_text(
    aes(label = round(mean_hr, 1), y = mean_hr + se + 0.8),
    size = 3.2,
    fontface = "bold",
    colour = "grey20"
  ) +
  scale_fill_manual(values = bar_colours) +
  scale_y_continuous(
    breaks = seq(60, 90, 10),
    expand = expansion(mult = c(0, 0.05))
  ) +
  coord_cartesian(ylim = c(60, 92)) +
  labs(
    title    = "Mean heart rate by UTCI thermal stress class",
    subtitle = "Error bars = ±1 SE · n = 189,329 observations with valid HR",
    x = NULL,
    y = "HR (bpm)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, colour = "grey40", size = 10),
    axis.text.y = element_text(colour = "grey40"),
    axis.title.y = element_text(colour = "grey40", size = 10),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.4),
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(colour = "grey50", size = 10)
  )
```


```{r hr-by-class, fig.width=5, fig.height=5}
phr
ggsave("../../outputs/climate_healt/phr_bars.png", phr, width = 7, height = 7, dpi = 300)
```

## 4. Blood pressure — U-shaped curve

```{r}
library(tidyverse)
library(scales)

utci_levels <- c(
  "V.strong cold", "Strong cold", "Moderate cold", "Slight cold",
  "No stress",
  "Moderate heat", "Strong heat", "V.strong heat"
)

bp_class <- df |>
  mutate(
    utci_class_short = factor(utci_class_short, levels = utci_levels)
  ) |>
  group_by(utci_class_short) |>
  summarise(
    Systolic  = mean(bphigh, na.rm = TRUE),
    Diastolic = mean(bplow,  na.rm = TRUE),
    n_systolic  = sum(!is.na(bphigh)),
    n_diastolic = sum(!is.na(bplow)),
    .groups = "drop"
  ) |>
  mutate(x = as.numeric(utci_class_short)) |>
  pivot_longer(
    c(Systolic, Diastolic),
    names_to = "measure",
    values_to = "mmhg"
  ) |>
  mutate(
    n = if_else(measure == "Systolic", n_systolic, n_diastolic)
  )
```

```{r}
format_p <- function(p) {
  case_when(
    is.na(p) ~ NA_character_,
    p == 0 | p < 2.2e-16 ~ "p < 2.2e-16",
    TRUE ~ paste0("p = ", formatC(p, format = "e", digits = 2))
  )
}

bp_test <- function(data, bp_var, measure_label) {
  
  neutral <- data |>
    filter(utci_stress_class == "No thermal stress") |>
    pull({{ bp_var }}) |>
    na.omit()
  
  cold_extreme <- data |>
    filter(
      utci_stress_class %in% c(
        "Strong cold stress",
        "Very strong cold stress"
      )
    ) |>
    pull({{ bp_var }}) |>
    na.omit()
  
  heat_extreme <- data |>
    filter(
      utci_stress_class %in% c(
        "Strong heat stress",
        "Very strong heat stress"
      )
    ) |>
    pull({{ bp_var }}) |>
    na.omit()
  
  w_cold <- wilcox.test(cold_extreme, neutral, exact = FALSE)
  w_heat <- wilcox.test(heat_extreme, neutral, exact = FALSE)
  
  group_mean <- c(
    mean(cold_extreme, na.rm = TRUE),
    mean(heat_extreme, na.rm = TRUE)
  )
  
  neutral_mean <- mean(neutral, na.rm = TRUE)
  
  tibble(
    measure = measure_label,
    comparison_type = c("cold", "heat"),
    comparison = c(
      "strong/very strong cold vs neutral",
      "strong/very strong heat vs neutral"
    ),
    mean_group = group_mean,
    mean_neutral = neutral_mean,
    delta = group_mean - neutral_mean,
    p_value = c(w_cold$p.value, w_heat$p.value),
    p_label = format_p(p_value),
    n_group = c(length(cold_extreme), length(heat_extreme)),
    n_neutral = length(neutral)
  )
}

bp_tests <- bind_rows(
  bp_test(df, bphigh, "Systolic"),
  bp_test(df, bplow,  "Diastolic")
) |>
  mutate(
    across(c(mean_group, mean_neutral, delta), ~ round(.x, 1))
  )
```

```{r}
neutral_lines <- bp_tests |>
  group_by(measure) |>
  summarise(
    mean_neutral = first(mean_neutral),
    .groups = "drop"
  )

caption_text <- bp_tests |>
  mutate(
    delta_label = paste0(if_else(delta >= 0, "+", ""), delta, " mmHg"),
    comparison_label = recode(
      comparison_type,
      cold = "Cold extremes",
      heat = "Heat extremes"
    ),
    text = paste0(
      measure, ": Δ ", delta_label, ", ", p_label
    )
  ) |>
  group_by(comparison_label) |>
  summarise(
    text = paste(text, collapse = " · "),
    .groups = "drop"
  ) |>
  mutate(
    text = paste0(comparison_label, " — ", text)
  ) |>
  pull(text) |>
  paste(collapse = "\n")

label_df <- bp_class |>
  filter(
    utci_class_short %in% c(
      "V.strong cold",
      "Strong cold",
      "No stress",
      "Strong heat",
      "V.strong heat"
    )
  ) |>
  mutate(
    label = round(mmhg, 1),
    y_label = case_when(
      measure == "Systolic"  ~ mmhg + 1.0,
      measure == "Diastolic" ~ mmhg - 1.1
    )
  )

pbp <- ggplot(bp_class, aes(x = x, y = mmhg)) +
  annotate(
    "rect",
    xmin = 0.5, xmax = 2.5,
    ymin = -Inf, ymax = Inf,
    fill = "#dbeafe",
    alpha = 0.45
  ) +
  annotate(
    "rect",
    xmin = 4.5, xmax = 5.5,
    ymin = -Inf, ymax = Inf,
    fill = "#fef3c7",
    alpha = 0.55
  ) +
  annotate(
    "rect",
    xmin = 6.5, xmax = 8.5,
    ymin = -Inf, ymax = Inf,
    fill = "#fee2e2",
    alpha = 0.45
  ) +
  geom_hline(
    data = neutral_lines,
    aes(yintercept = mean_neutral, colour = measure),
    inherit.aes = FALSE,
    linetype = "dashed",
    linewidth = 0.45,
    alpha = 0.55,
    show.legend = FALSE
  ) +
  geom_line(
    aes(colour = measure, group = measure, linewidth = measure),
    lineend = "round"
  ) +
  geom_point(
    aes(fill = measure),
    shape = 21,
    size = 3.8,
    colour = "white",
    stroke = 0.9
  ) +
  geom_text(
    data = label_df,
    aes(
      y = y_label,
      label = label,
      colour = measure
    ),
    size = 3,
    fontface = "bold",
    show.legend = FALSE
  ) +
  scale_x_continuous(
    breaks = seq_along(utci_levels),
    labels = utci_levels,
    expand = expansion(mult = c(0.03, 0.03))
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 6),
    expand = expansion(mult = c(0.08, 0.08))
  ) +
  scale_colour_manual(
    values = c(
      "Systolic" = "#ef4444",
      "Diastolic" = "#3b82f6"
    )
  ) +
  scale_fill_manual(
    values = c(
      "Systolic" = "#ef4444",
      "Diastolic" = "#3b82f6"
    )
  ) +
  scale_linewidth_manual(
    values = c(
      "Systolic" = 1.35,
      "Diastolic" = 1.05
    )
  ) +
  labs(
    title = "Blood pressure is higher at thermal extremes",
    subtitle = "Mean systolic and diastolic blood pressure by UTCI class.",
    caption = caption_text,
    x = NULL,
    y = "Blood pressure (mmHg)",
    colour = NULL,
    fill = NULL,
    linewidth = NULL
  ) +
  guides(
    linewidth = "none",
    colour = guide_legend(
      override.aes = list(linewidth = 1.4, size = 3)
    )
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "top",
    legend.justification = "left",
    legend.text = element_text(size = 10, colour = "grey35"),
    axis.text.x = element_text(angle = 35, hjust = 1, colour = "grey35", size = 10),
    axis.text.y = element_text(colour = "grey45"),
    axis.title.y = element_text(colour = "grey35", size = 10),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.35),
    plot.title = element_text(face = "bold", size = 14, colour = "grey15"),
    plot.subtitle = element_text(size = 10.2, colour = "grey45"),
    plot.caption = element_text(
      size = 8.6,
      colour = "grey45",
      hjust = 0,
      lineheight = 1.05,
      margin = margin(t = 8)
    )
  )
```

```{r hr-by-class, fig.width=7, fig.height=7}
pbp
ggsave("../../outputs/climate_healt/pbp_mean.png", pbp, width = 7, height = 7, dpi = 300)

```



## 7. Physical activity — inverted-J curve

```{r steps-j}
library(tidyverse)
library(scales)

utci_levels <- c(
  "V.strong cold", "Strong cold", "Moderate cold", "Slight cold",
  "No stress",
  "Moderate heat", "Strong heat", "V.strong heat"
)

utci_colours <- c(
  "V.strong cold" = "#1d4ed8",
  "Strong cold"   = "#3b82f6",
  "Moderate cold" = "#93c5fd",
  "Slight cold"   = "#bfdbfe",
  "No stress"     = "#fde68a",
  "Moderate heat" = "#fbbf24",
  "Strong heat"   = "#f97316",
  "V.strong heat" = "#ef4444"
)

steps_class <- df |>
  mutate(
    utci_class_short = factor(utci_class_short, levels = utci_levels)
  ) |>
  group_by(utci_class_short) |>
  summarise(
    mean_steps = mean(steps, na.rm = TRUE),
    sd_steps   = sd(steps, na.rm = TRUE),
    n          = sum(!is.na(steps)),
    se_steps   = sd_steps / sqrt(n),
    .groups    = "drop"
  )

neutral_steps <- steps_class |>
  filter(utci_class_short == "No stress") |>
  pull(mean_steps)

steps_class <- steps_class |>
  mutate(
    pct_neutral = mean_steps / neutral_steps,
    label_pct = paste0(round(pct_neutral * 100), "%"),
    label_steps = number(mean_steps, accuracy = 1)
  )

ggplot(steps_class, aes(x = utci_class_short, y = mean_steps)) +
  annotate(
    "rect",
    xmin = 0.5, xmax = 2.5,
    ymin = -Inf, ymax = Inf,
    fill = "#dbeafe",
    alpha = 0.35
  ) +
  annotate(
    "rect",
    xmin = 4.5, xmax = 5.5,
    ymin = -Inf, ymax = Inf,
    fill = "#fef3c7",
    alpha = 0.55
  ) +
  annotate(
    "rect",
    xmin = 6.5, xmax = 8.5,
    ymin = -Inf, ymax = Inf,
    fill = "#fee2e2",
    alpha = 0.35
  ) +
  geom_hline(
    yintercept = neutral_steps,
    linetype = "dashed",
    linewidth = 0.45,
    colour = "grey45",
    alpha = 0.65
  ) +
  geom_col(
    aes(fill = utci_class_short),
    width = 0.68,
    alpha = 0.88,
    show.legend = FALSE
  ) +
  geom_errorbar(
    aes(ymin = mean_steps - se_steps, ymax = mean_steps + se_steps),
    width = 0.18,
    linewidth = 0.45,
    colour = "grey35"
  ) +
  geom_line(
    aes(group = 1),
    linewidth = 0.85,
    colour = "grey25",
    alpha = 0.65
  ) +
  geom_point(
    size = 2.8,
    shape = 21,
    fill = "white",
    colour = "grey20",
    stroke = 0.8
  ) +
  geom_text(
    aes(
      label = label_pct,
      y = mean_steps + se_steps + max(mean_steps, na.rm = TRUE) * 0.045
    ),
    size = 3.2,
    fontface = "bold",
    colour = "grey25"
  ) +
  scale_fill_manual(values = utci_colours) +
  scale_y_continuous(
    labels = number_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.14))
  ) +
  labs(
    title = "Physical activity decreases under thermal stress",
    subtitle = "Mean hourly steps by UTCI class. Labels show percentage relative to neutral conditions.",
    x = NULL,
    y = "Steps per hour",
    caption = "Error bars represent ±1 SE. Dashed line indicates the mean value under neutral thermal conditions."
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 35, hjust = 1, colour = "grey35", size = 10),
    axis.text.y = element_text(colour = "grey45"),
    axis.title.y = element_text(colour = "grey35", size = 10),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.35),
    plot.title = element_text(face = "bold", size = 14, colour = "grey15"),
    plot.subtitle = element_text(size = 10.2, colour = "grey45"),
    plot.caption = element_text(size = 8.8, colour = "grey45", hjust = 0),
    plot.margin = margin(8, 12, 8, 8)
  )
```

---

## 8. Circadian rhythm of HR

```{r circadian}
hr_hour <- df |>
  group_by(hour) |>
  summarise(
    hr_mean = mean(heartrate, na.rm = TRUE),
    hr_se   = sd(heartrate, na.rm = TRUE) / sqrt(sum(!is.na(heartrate))),
    .groups = "drop"
  )

ggplot(hr_hour, aes(hour, hr_mean)) +
  geom_ribbon(aes(ymin = hr_mean - hr_se, ymax = hr_mean + hr_se),
              fill = "#3b82f6", alpha = 0.2) +
  geom_line(colour = "#3b82f6", linewidth = 1.2) +
  geom_point(colour = "#1d4ed8", size = 2) +
  scale_x_continuous(breaks = seq(0, 23, 3),
                     labels = paste0(seq(0, 23, 3), ":00")) +
  labs(
    title    = "Circadian rhythm of heart rate",
    subtitle = "Nadir ~04:00 (60 bpm) · daytime plateau ~81 bpm · averaged across all users",
    x = "Hour of day", y = "HR (bpm)"
  ) +
  theme_minimal(base_size = 12)
```

---

## 9. Summary table

```{r summary-table}
class_summary <- df |>
  group_by(utci_class_short) |>
  summarise(
    n_obs      = n(),
    utci_mean  = mean(utci_c,     na.rm = TRUE),
    hr         = mean(heartrate,  na.rm = TRUE),
    systolic   = mean(bphigh,     na.rm = TRUE),
    diastolic  = mean(bplow,      na.rm = TRUE),
    steps_h    = mean(steps,      na.rm = TRUE),
    breathrate = mean(breathrate, na.rm = TRUE),
    skintemp   = mean(skintemp,   na.rm = TRUE),
    .groups    = "drop"
  ) |>
  mutate(across(where(is.numeric), ~ round(.x, 1)))

knitr::kable(
  class_summary,
  col.names = c("UTCI class", "N obs", "UTCI °C", "HR bpm",
                "Systolic mmHg", "Diastolic mmHg",
                "Steps/h", "Breath rate", "Skin temp °C"),
  caption   = "Mean physiological values by UTCI thermal stress class"
)
```

---

## 10. Methodological notes

**Missing values** — physiological variables are missing from 28% (HR) to 92% (sleeprate). Each analysis uses the available subset for that variable; there is no complete-case dataset. Results for extreme classes (very strong cold/heat) are based on small n (~76–796 obs) and should be interpreted cautiously.

**Circadian confounding** — UTCI is higher during the day, when HR is naturally elevated. The mixed model (section 4) controls for this via `hour_sin`/`hour_cos`. The by-class analyses (sections 3 and 6) do not explicitly control for time of day.

**Outdoor UTCI vs actual exposure** — UTCI values come from a weather station, not the wearable. Users spending time indoors in air-conditioned spaces are not actually exposed to the recorded UTCI.

**Interpreting lmer coefficients** — the UTCI_C coefficient (~0.14 bpm/°C) is the mean within-user effect, adjusted for hour and city. The raw between-class effect (section 3) is larger because it absorbs seasonal and geographic differences.

```{r session}
sessionInfo()
```
