---
title: "Particle measurements quality control"
author: "Alessandro Fuschi"
editor: source
---

## Setup

```{r, message=FALSE}
library(tidyverse)

cfg <- list(
  notebook_id = "fixing_particle_vars",
  data_dir = file.path("data", "5min"),
  output_dir = file.path("outputs", "fixing_particle_vars"),
  timezone = "UTC"
)

dir.create(cfg$output_dir, recursive = TRUE, showWarnings = FALSE)
```

## Data import

The five-minute MyAir table is imported using `userId` and `bucket_5min` as the analytical key.

Rows belonging to duplicated analytical keys are removed completely, because their origin cannot be resolved unambiguously at this stage.

```{r, message=FALSE}
read_5min <- \(filename) {
  read_csv(
    file.path(cfg$data_dir, filename),
    na = c("", "NA", "NULL"),
    locale = locale(tz = cfg$timezone),
    col_types = cols(
      userId = col_character(),
      bucket_5min = col_datetime(format = "%Y-%m-%d %H:%M:%S")
    ),
    show_col_types = FALSE
  ) %>%
    add_count(userId, bucket_5min, name = "key_n") %>%
    filter(key_n == 1) %>%
    select(-key_n)
}
```

```{r}
air <- read_5min("myair_5min.csv.gz")

rm(read_5min)
```

## Variable selection

Mean particulate-mass and particle-count measurements are retained for each five-minute interval.

Temperature and humidity minima and maxima are also retained for subsequent calculation of within-bucket variability.

```{r}
air <- air %>%
  transmute(
    userId,
    bucket_5min,

    temperature = temperature_mean,
    temperature_min,
    temperature_max,

    humidity = humidity_mean,
    humidity_min,
    humidity_max,

    pm1 = pm1_mean,
    pm25 = pm25_mean,
    pm10 = pm10_mean,

    pc03 = pc03_mean,
    pc05 = pc05_mean,
    pc1 = pc1_mean,
    pc25 = pc25_mean,
    pc5 = pc5_mean,
    pc10 = pc10_mean,

    uvb = uvb_mean,
    light = light_mean
  )
```

## Particle measurements quality control

Particle-count variables represent cumulative counts above increasing particle-size thresholds. They are therefore expected to follow the monotonic relation:

$$
\mathrm{PC}_{0.3} \geq
\mathrm{PC}_{0.5} \geq
\mathrm{PC}_{1.0} \geq
\mathrm{PC}_{2.5} \geq
\mathrm{PC}_{5.0} \geq
\mathrm{PC}_{10}
$$

A five-minute particle block is classified as invalid when:

- the cumulative particle-count ordering is violated;
- at least one particulate-mass value exceeds 1,000 µg/m³.

Because particulate-mass and particle-count variables originate from the same sensor block, all PM and PC measurements are set to missing when either condition is detected.

```{r}
air <- air %>%
  mutate(
    invalid_particle_order = coalesce(
      pc03 < pc05 |
        pc05 < pc1 |
        pc1 < pc25 |
        pc25 < pc5 |
        pc5 < pc10,
      FALSE
    ),

    invalid_particle_mass = coalesce(
      pm1 > 1000 |
        pm25 > 1000 |
        pm10 > 1000,
      FALSE
    ),

    invalid_particles =
      invalid_particle_order |
      invalid_particle_mass,

    pm1 = if_else(invalid_particles, NA_real_, pm1),
    pm25 = if_else(invalid_particles, NA_real_, pm25),
    pm10 = if_else(invalid_particles, NA_real_, pm10),

    pc03 = if_else(invalid_particles, NA_real_, pc03),
    pc05 = if_else(invalid_particles, NA_real_, pc05),
    pc1 = if_else(invalid_particles, NA_real_, pc1),
    pc25 = if_else(invalid_particles, NA_real_, pc25),
    pc5 = if_else(invalid_particles, NA_real_, pc5),
    pc10 = if_else(invalid_particles, NA_real_, pc10)
  )
```

## Quality-control summary

```{r}
air %>%
  count(
    invalid_particle_order,
    invalid_particle_mass,
    invalid_particles
  ) %>%
  mutate(percent = 100 * n / sum(n)) %>%
  knitr::kable(
    format = "pipe",
    digits = 2,
    col.names = c(
      "Invalid count ordering",
      "Invalid particulate mass",
      "Invalid particle block",
      "Observations",
      "Percent"
    ),
    caption = "Particle quality-control classification"
  )
```














